---
title: "Cell line selection"
output: html_notebook
---

!!!!!!!!!!!!This script was used for the initial selection and is not used anymore!!!!!!!!!!!!!!!!!!!!
Based on the species that were selected to be included in the illumina or silencer projects we have to select which cell lines to use in case there is more than one.

```{r}
library(tidyverse)
library(cowplot)
library(readxl)
library(googlesheets4)

theme_set(theme_classic())

here::here()

path<-here::here()
```

# read species selection

```{r}
selection <- read_sheet(
  "https://docs.google.com/spreadsheets/d/1k-0sFR4AuaA0hAM2qeUZ8IMvLx96vY2JIQbwrNQtLj0/edit?usp=drive_link"
) %>%
  arrange(-silencer, -illumina, species) %>%
  select(silencer, illumina, species)

species_list <- readRDS(paste0(path, "/output/Species_List_with_genome_and_cl_info2.rds")) %>%
  full_join(selection) %>%
  mutate(
    illumina = if_else(is.na(illumina), F, illumina),
    silencer = if_else(is.na(silencer), F, silencer),
    tetrapod = !illumina & superclass != "Fish",
    selected = illumina | silencer | tetrapod
  ) %>%
  select(
    "scientific_name",
    "species",
    "iucn_category",
    "individuals",
    "taxon_id",
    "genus" ,
    "family",
    "order",
    "class",
    "superclass",
    "contig_n50",
    "assembly_level",
    "scaffold_n50",
    "genome_size",
    "gene_count",
    "genome_match",
    "species_genome",
    "fibro_l",
    "cat",
    "illumina",
    "silencer",
    "tetrapod",
    "selected"
  )


sample_table <- readRDS(paste0(path, "/output/Sample_List_all_species_clean.rds")) %>%
  mutate(cryotube = animal_ID %in% cryo$animal_ID) %>%
  left_join(select(species_list, species, illumina, silencer, tetrapod, selected)) %>%
  select(animal_ID, 1:11, 13:ncol(.)) %>%
  left_join(rownames_to_column(as.data.frame(summary), var = "animal_ID")) %>%
  arrange(-silencer, -illumina, -tetrapod, -selected)

cryo <- readRDS(paste0(path, "/data/Cryo_stocks_summary.rds")) %>%
  group_by(animal_ID) %>%
  filter(type2 == "Fibroblast") %>%
  filter(CT_total > 0) %>%
  filter(!is.na(CT_total)) %>%
  select(animal_ID, n_passages, max_passage, CT_total)
```


```{r}
cryzoo_seq <- read_sheet(
  "https://docs.google.com/spreadsheets/d/1YI-0bB7OFfjaUM4af9Lv3BAd1X0x9CoCTKZN3BQM4KA/edit?usp=sharing"
)

ggplot(cryzoo_seq) +
  geom_tile(aes(paste(species, animal_ID), type, fill = processed_by), col =
              "black") +
  coord_flip() +
  xlab("")

summary <- cryzoo_seq %>%
  group_by(animal_ID, species, type) %>%
  summarize(sum = n()) %>%
  pivot_wider(
    id_cols = c(animal_ID),
    names_from = type,
    values_from = sum
  )

summary[is.na(summary)] <- 0
colnames(summary)[3:4]<-paste0(colnames(summary)[3:4],"-seq")

summary_species <- cryzoo_seq %>%
  group_by(species, type) %>%
  summarize(sum = n()) %>%
  pivot_wider(
    id_cols = c(species),
    names_from = type,
    values_from = sum
  )

summary_species[is.na(summary_species)] <- 0


#write_sheet(data = summary_species, ss = "https://docs.google.com/spreadsheets/d/1YI-0bB7OFfjaUM4af9Lv3BAd1X0x9CoCTKZN3BQM4KA/edit?usp=sharing", sheet = "Sequencing_summary_per_Species")

summary<-summary %>% 
  column_to_rownames(var = "animal_ID")

summary<-summary>0

```

check which sample have RNA extracted already or at least have a pellet with cells ready for extraction

```{r}
RNA <- read_sheet(
  "https://docs.google.com/spreadsheets/d/1SRid9RGRfcCpGAWsAXG8h0wE-7p1fhOtwvkMJgR1s8w/edit?usp=sharing",
  sheet = "RNA Extractions",
  .name_repair = "universal"
) %>%
  rename(animal_ID = Cryozoo.Number,
         Volume_RNA = ul,
         c_RNA = Ribonucleic.Acid.ng.µl.) %>%
  select(c(animal_ID, Volume_RNA, c_RNA, Bioanalyzer.result..RIN.))


pellet <- read_sheet(
  "https://docs.google.com/spreadsheets/d/1SRid9RGRfcCpGAWsAXG8h0wE-7p1fhOtwvkMJgR1s8w/edit?usp=drive_link",
  sheet = "Pelet recollit per DNA i RNA",
  col_names = c("animal_ID", "species", "vials_collected", "DNA_ex", "RNA_ex"),
  col_types = c("ccccc")
) %>%
  mutate(
    pellet = case_when(
      str_detect(vials_collected, pattern = "\\s*(RNA|DNA)") ~ str_count(vials_collected, pattern =
                                                                           "\\s*(RNA|DNA)"),
      is.na(vials_collected) ~ 0,
      .default = as.numeric(vials_collected)
    ),
    pellet = if_else(is.na(pellet), 1, pellet),
    pellet_type = case_when(
      str_detect(vials_collected, pattern = "\\s*(RNA&DNA)") ~ "both",
      str_detect(vials_collected, pattern = "\\s*(DNA)") ~
        "DNA",
      str_detect(vials_collected, pattern = "\\s*(RNA)") ~
        "RNA",
      pellet == 0 ~ NA_character_,
      T ~ "DNA"
    )
  ) %>%
  left_join(RNA) %>%
  mutate(DNA_ex = !is.na(DNA_ex),
         RNA_ex = !is.na(RNA_ex) | !is.na(c_RNA)) %>%
  separate(
    animal_ID,
    remove = F,
    sep = "[^[:alnum:]]+",
    into = c("animal_ID1", "animal_ID2")
  ) %>%
  mutate(animal_ID = if_else(
    as.numeric(str_remove(animal_ID1, "\\D")) > as.numeric(str_remove(animal_ID2, "\\D")) &
      !is.na(animal_ID2),
    animal_ID2,
    animal_ID1
  )) %>%
  mutate(animal_ID2 = if_else(animal_ID == animal_ID1, animal_ID2, animal_ID1)) %>%
  group_by(animal_ID, animal_ID2) %>%
  summarise(
    DNA_ex = sum(DNA_ex) > 0,
    RNA_ex = sum(RNA_ex) > 0,
    pellet = sum(pellet),
    pellet_type = paste(pellet_type, collapse = ",")
  ) %>%
  separate(pellet_type, sep = ",", c("pellet_type1", "pellet_type2")) %>%
  ungroup() %>%
  mutate(
    pellet_type = case_when(
      pellet_type1 == "RNA" & pellet_type2 == "RNA" ~ "RNA",
      pellet_type1 == "DNA" & pellet_type2 == "DNA" ~ "DNA",
      pellet_type1 == "RNA" & pellet_type2 == "DNA" ~ "both",
      pellet_type1 == "DNA" & pellet_type2 == "RNA" ~ "both",
      pellet_type2 == "NA" ~ NA_character_,
      pellet_type1 == "NA" ~ pellet_type2,
      T ~ pellet_type1
    )
  ) %>%
  select(-c(pellet_type1, pellet_type2))

```

```{r}
selected_samples <- sample_table %>%
  filter(cell_type == "Fibro") %>%
  filter(illumina | silencer) %>%
  left_join(pellet) %>%
  left_join(cryo) %>%
  mutate(
    from_cryo = CT_total > 4 & n_passages > 2,
    RNA_ex = if_else(is.na(RNA_ex), F, RNA_ex),
    DNA_ex = if_else(is.na(DNA_ex), F, DNA_ex),
    pellet = if_else(is.na(pellet), 0, pellet),
    pellet_RNA = pellet > 0 & pellet_type %in% c("RNA", "both")
  ) %>%
  group_by(species) %>%
  mutate(
    selected = case_when(
      is.na(CT_total) ~ F,
      RNA_ex ~ T,
      DNA_ex ~ T,
      pellet_RNA ~ T,
      seq ~ T,
      from_cryo ~ T,
      n() < 3 ~ T,
      T ~ F
    ),
    n_samples = n()
  ) %>%
  arrange(-selected) %>%
  mutate(rep_type = if_else(n_samples > 1, "biorep", "techrep"),
         rep = row_number()) %>%
  arrange(species, rep)


#write_excel_csv(selected_samples,paste0(path,"/data/selected_celllines_RNA.csv"))
```




```{r}
selected_samples_short <- selected_samples %>%
  mutate(species_in = if_else(silencer, "silencer", "illumina")) %>%
  select(
    c(
      animal_ID,
      sample_id,
      species,
      superclass,
      seq,
      DNA_ex,
      RNA_ex,
      pellet,
      CT_total,
      n_passages,
      rep_type,
      rep,
      species_in,
      selected
    )
  ) %>%
  arrange(species, -selected)

#write_sheet(species_list, ss = "https://docs.google.com/spreadsheets/d/1csFMpLFtTr1_-hF1Nncduzjl9uIjKHRh0w7kntRZM2A/edit?usp=drive_link", sheet = "Species")

#write_sheet(selected_samples_short, ss = "https://docs.google.com/spreadsheets/d/1csFMpLFtTr1_-hF1Nncduzjl9uIjKHRh0w7kntRZM2A/edit?usp=drive_link", sheet = "Cell_lines")

#write_sheet(selected_samples, ss = "https://docs.google.com/spreadsheets/d/1csFMpLFtTr1_-hF1Nncduzjl9uIjKHRh0w7kntRZM2A/edit?usp=drive_link", sheet = "Cell_lines_full")

#write_sheet(sample_table, ss = "https://docs.google.com/spreadsheets/d/1csFMpLFtTr1_-hF1Nncduzjl9uIjKHRh0w7kntRZM2A/edit?usp=drive_link", sheet = "All_samples")


selected_samples2 <- subset(selected_samples, selected)

table(selected_samples2$sex)
table(selected_samples2$origin)

hist(table(selected_samples2$species))



selected_samples_full <- bind_rows(selected_samples2,
                                   selected_samples2[which(selected_samples2$reps == "techrep1"), ] %>%
                                     mutate(reps = "techrep2"))

hist(table(selected_samples_full$species))

```



# check which of the already extracted RNAs to use

```{r}

# read_sheet("https://docs.google.com/spreadsheets/d/1y4QBygDAAi7GRdSAZ9N4S1k7vz2r0SjkLIdhfUAZ1XY/edit?usp=sharing",range = "Extractions") %>% 
#  mutate(project2=if_else(animal_ID%in%selected_samples$animal_ID|animal_ID%in%selected_samples$animal_ID2,"illumina","cryozoo")) %>% write_sheet("https://docs.google.com/spreadsheets/d/1y4QBygDAAi7GRdSAZ9N4S1k7vz2r0SjkLIdhfUAZ1XY/edit?usp=sharing",sheet =  "Extractions")

```


# Cell lines harvesting

```{r}


harvesting <- read_sheet(
  "https://docs.google.com/spreadsheets/d/1vyBZwssG8ndUcGmDg4kLEPwOJJk15bOrXztO876HCOs/edit?usp=sharing",
  "tracking",
  col_types = "nccccnccccDlDnccccnlllnnnDc"
)

summary <- harvesting %>%
  group_by(species) %>%
  summarize(
    vials_tested = n(),
    lines_tested = length(unique(animal_ID)),
    lines_harvested = sum(Harvested, na.rm = T),
    DNA = sum(DNA, na.rm = T),
    RNA = sum(RNA, na.rm = T),
    ATAC = sum(ATAC, na.rm = T),
    lines = paste(unique(animal_ID), collapse = ",")
  )

# write_sheet(
#   summary,
#   "https://docs.google.com/spreadsheets/d/1vyBZwssG8ndUcGmDg4kLEPwOJJk15bOrXztO876HCOs/edit?usp=sharing",
#   "summary"
# )

missing_species <- setdiff(subset(selection, silencer)$species, summary$species)

missing_species


# new_lines <- summary %>%
#   filter(lines_harvested < 2) %>%
#   select(species, lines_tested) %>%
#   left_join(selected_samples) %>%
#   filter(!(animal_ID %in% harvesting$animal_ID)) %>%
#   filter(!is.na(animal_ID)) %>%
#   
#   write_sheet(
#     "https://docs.google.com/spreadsheets/d/1vyBZwssG8ndUcGmDg4kLEPwOJJk15bOrXztO876HCOs/edit?usp=sharing",
#     "new_lines"
#   )

# harvesting %>%
#   group_by(animal_ID, species) %>%
#   summarize(tries = n(), Harvested = sum(Harvested) > 0) %>%
#   filter(tries < 2 & !Harvested) %>%
#   left_join(selected_samples) %>%
#   filter(CT_total > 2) %>%
#   
#   
#   write_sheet(
#     "https://docs.google.com/spreadsheets/d/1vyBZwssG8ndUcGmDg4kLEPwOJJk15bOrXztO876HCOs/edit?usp=sharing",
#     "repeat"
#   )

```




```{r}


plot_df <- selected_samples %>%
  mutate(ispellet = pellet > 0) %>%
  select(species,
         animal_ID,
         RNA_ex,
         from_cryo,
         silencer,
         illumina,
         pellet_RNA) %>%
  rownames_to_column(var = "index") %>%
  pivot_longer(
    cols = c(pellet_RNA, RNA_ex, from_cryo),
    names_to = "type",
    values_to = "value"
  ) %>%
  mutate(
    index = as.numeric(index),
    quarter = case_when(
      index < nrow(selected_samples) / 4 ~ "top",
      index < nrow(selected_samples) / 2 ~ "middle1",
      index > nrow(selected_samples) / 2 &
        index < (nrow(selected_samples) / 4) * 3 ~ "middle2",
      T ~ "bottom"
    ),
    type = factor(type, levels = rev(c(
      "from_cryo", "pellet_RNA", "RNA_ex"
    )))
  )

table(plot_df$quarter)


plot_df[is.na(plot_df)] <- F

ggplot(plot_df) +
  geom_tile(aes(
    x = type,
    y = reorder(animal_ID, value),
    fill = value
  ), col = "white") +
  facet_wrap( ~ silencer, ncol = 4, scales = "free") +
  theme(
    strip.text = element_blank(),
    strip.background = element_blank(),
    axis.text.x = element_text(angle = 90, vjust = 0.3)
  ) +
  ggsci::scale_fill_npg() +
  xlab("")


table(plot_df$type, plot_df$value)
```

## Find Cryo tubes

```{r}



extr_long <- readRDS(paste0(path, "/data/Extractions_long.rds"))


IDs <- c(
  "A495
A524
A345
A78
A514
A50
A469
A15
A542
A182
A537
A357
A8
A260
A476
A544
A37
A529
A422
A65
A434
A322
A511
A289
A262
A424
A102
A489
A533
A472
A247
A446
A493
A81
A77
A462
A274
A532
A157
A439
A522
A353
A73
A67
A92
A526
A403
A374
A288
A509
A70
A278
A312
A286
A482
A558
A392
A422
A391
A81
A461
A308
A460
A394
A11
A13"
) %>%
  str_split(pattern = "\n", simplify = F) %>%
  unlist()

positions <- extr_long %>%
  filter(type == "RNA") %>%
  mutate(column = as.numeric(str_extract(column, "\\d+"))) %>%
  mutate(row = as.numeric(str_extract(row, "\\d+"))) %>%
  mutate(animal_ID = str_extract(cryotube, pattern = "[Aa]\\d+")) %>%
  filter(animal_ID %in% IDs)

ggplot(positions) +
  geom_raster(aes(column, row)) +
  geom_hline(yintercept = 0.5:10.5) +
  geom_vline(xintercept = 0.5:10.5) +
  scale_y_reverse(breaks = 1:10) +
  scale_x_continuous(breaks = 1:12, position = "top") +
  geom_text(aes(column, row, label = cryotube),
            col = "white",
            size = 2) +
  facet_wrap( ~ box)



#write_excel_csv(positions_pellet,paste0(path,"/output/positions_pellet.csv"))
```

