---
title: "Cell line selection - ongoing"
output: html_notebook
---

 !!!!! Archived not in use anymore !!!!!

Based on the species that were selected to be included in the illumina or silencer projects we have to select which cell lines to use this has to be repeated once in a while to incorporate new samples. The main sources of information are 1. What has been sequenced and 2. What has been harvested

```{r}
library(tidyverse)
library(cowplot)
library(readxl)
library(googlesheets4)
library(stringr)

theme_set(theme_classic())

here::here()

path<-here::here()
```

# read species selection

```{r}
selection <- read_sheet(
  "https://docs.google.com/spreadsheets/d/1k-0sFR4AuaA0hAM2qeUZ8IMvLx96vY2JIQbwrNQtLj0/edit?usp=drive_link"
) %>%
  arrange(-silencer, -illumina, species) %>%
  mutate(taxon_id=as.character(taxon_id)) %>% 
  select(taxon_id,silencer, illumina, species) 
```

Join with the table containing all species and select relevant columns

```{r}
species_list <- readRDS(paste0(path, "/output/Species_List_with_genome_and_cl_info.rds"))  %>%
  select(
    "scientific_name",
    "species",
    "iucn_category",
    "individuals",
    "taxon_id",
    "genus" ,
    "family",
    "order",
    "class",
    "superclass",
    "contig_n50",
    "scaffold_n50",
    "assembly_level",
    "genome_size",
    "gene_count",
    "genome_match",
    "species_genome",
    "TOGA",
    "fibro_l",
    "cat",
    "illumina",
    "silencer",
    "tetrapod",
    "selected",
    "project",
    "i_psc",
    "cell_line"
  )

```

# Do we have cryostocks?

```{r}
cryo <- readRDS(paste0(path, "/data/Cryo_stocks_summary.rds")) %>%
  group_by(animal_id) %>%
  filter(type2 == "Fibroblast") %>%
  filter(CT_total > 0) %>%
  filter(!is.na(CT_total)) %>%
  select(animal_id, n_passages, max_passage, CT_total)
```

# What has been sequenced already?

## Summarized per cell line

This table is generated in CGL_ZOO_Project_status_overview.Rmd
```{r}
cryzoo_seq <- read_sheet(
  "https://docs.google.com/spreadsheets/d/1YI-0bB7OFfjaUM4af9Lv3BAd1X0x9CoCTKZN3BQM4KA/edit?usp=sharing",
  range = "All_Sequenced"
)
```


```{r}

cz_plot_df<-cryzoo_seq %>% 
  #select(-scientific_name) %>% 
  left_join(select(species_list,-project))

ggplot(cz_plot_df) +
  geom_tile(aes(paste(species, animal_id), type), 
            col ="black") +
  coord_flip() +
  xlab("")
```


```{r}

ggplot(cz_plot_df) +
  geom_tile(aes(species, type), 
            col ="black") +
  coord_flip() +
  xlab("")

```

```{r}

class_col<-ggsci::pal_aaas()(length(unique(species_list$superclass)))
names(class_col)<-unique(species_list$superclass)

ggplot(cz_plot_df) +
  geom_bar(aes(type, fill = superclass) ) +
  coord_flip() +
  xlab("")+
  ylab("number of libraries")+
  scale_fill_manual(values=class_col)

```


```{r}
summary_sequenced <- cryzoo_seq %>%
  group_by(animal_id, species, type) %>%
  summarize(sum = n()) %>%
  pivot_wider(
    id_cols = c(animal_id),
    names_from = type,
    values_from = sum
  )

summary_sequenced[is.na(summary_sequenced)] <- 0

summary_sequenced<-summary_sequenced %>% 
  column_to_rownames(var = "animal_id")

summary_sequenced<-summary_sequenced>0

```

## Summarized per species
```{r}
summary_species <- cryzoo_seq %>%
  group_by(species, type) %>%
  summarize(sum = n()) %>%
  pivot_wider(
    id_cols = c(species),
    names_from = type,
    values_from = sum
  )

summary_species[is.na(summary_species)] <- 0

write_sheet(data = summary_species, 
            ss = "https://docs.google.com/spreadsheets/d/1YI-0bB7OFfjaUM4af9Lv3BAd1X0x9CoCTKZN3BQM4KA/edit?usp=sharing",
            sheet = "Sequencing_summary_per_Species")


```


# Which cell lines have been harvested already?

```{r}

harvesting <- read_sheet(
  "https://docs.google.com/spreadsheets/d/1vyBZwssG8ndUcGmDg4kLEPwOJJk15bOrXztO876HCOs/edit?usp=sharing",
  "tracking_silencer",
  col_types = "nccccncccclDlDnccccnlllnnnDc"
) %>% filter(!is.na(animal_id))

harvesting <-bind_rows(harvesting, read_sheet(
  "https://docs.google.com/spreadsheets/d/1vyBZwssG8ndUcGmDg4kLEPwOJJk15bOrXztO876HCOs/edit?usp=sharing",
  "tracking_all_illumina",
  col_types = "nccccncccclDlDnccccnlllnnnDc")
) %>% filter(!is.na(`cultured in`))

summary_harvest <- harvesting %>%
  group_by(species,Group) %>%
  summarize(
    vials_tested = n(),
    lines_tested = length(unique(animal_id)),
    lines_harvested = sum(Harvested, na.rm = T),
    DNA = sum(DNA, na.rm = T),
    RNA = sum(RNA, na.rm = T),
    ATAC = sum(ATAC, na.rm = T),
    lines = paste(unique(animal_id), collapse = ",")
  ) %>% 
  ungroup() %>% 
  arrange(Group) 

write_sheet(
  summary_harvest,
  "https://docs.google.com/spreadsheets/d/1vyBZwssG8ndUcGmDg4kLEPwOJJk15bOrXztO876HCOs/edit?usp=sharing",
  "summary"
)

harvested<-harvesting %>% 
  group_by(animal_id) %>% 
  summarize(harvested=sum(Harvested)>0) %>%
  mutate(harvested=case_when(harvested~"harvested",
                                !harvested ~"failed",
                                is.na(harvested)~"not_tested",
                                )) %>% 
  filter(!is.na(animal_id)) %>% 
  rename(animal_id=animal_id)

```


# Load table of samples

```{r}
sample_table <- readRDS(paste0(path, "/output/Sample_List_all_species_clean.rds")) %>%
  left_join(select(species_list, species,scientific_name, illumina, silencer, tetrapod, selected,project)) %>%
  select(all_of("animal_id"), everything()) %>% 
  left_join(rownames_to_column(as.data.frame(summary_sequenced), var = "animal_id")) %>%
  left_join(harvested) %>% 
  replace_na(list(WGS = FALSE, RNA_seq = FALSE,ATAC_seq = FALSE, WGS_Nanopore = FALSE)) %>% 
  arrange(-silencer, -illumina, -tetrapod, -selected) %>% 
  mutate(species=str_replace(species,pattern = "  ",replacement = " "))

```

# Cell line selection for "Molecular Cryozoo"
```{r}
selected_samples <- sample_table %>%
  filter(cell_type == "Fibro") %>% ## only samples ith Fibroblasts
  #filter(illumina | silencer) %>% ## samples that are in either the illumina or the silencer project
  left_join(cryo) %>%  ## have we harvested it already
  mutate(
    from_cryo = CT_total > 4 & n_passages > 2,  # do we have sufficient cryo tubes?
  ) %>% 
  left_join(harvested) %>% 
  group_by(species) %>%
  mutate(
    n_samples = n()
  ) %>%
  select(-selected) %>%
  mutate(rep_type = if_else(n_samples > 1, "biorep", "techrep"),
         rep = row_number(),
         selected=case_when(rep_type=="techrep"~T,
                            harvested=="harvested"~T,
                            seq|`WG-Seq`|`RNA-seq`|`ATAC-seq`|`WG-Nanopore`~T,
                            T~F)) %>%
  arrange(species, -selected, rep)




#write_excel_csv(selected_samples,paste0(path,"/data/selected_celllines_RNA.csv"))
```

## short version

```{r}

selected_samples_short <- selected_samples %>%
  mutate(species_in = if_else(silencer, "silencer", "illumina")) %>%
  select(
    c(
      animal_id,
      sample_id,
      species,
      superclass,
      seq,
      `RNA-seq`,
      `ATAC-seq`,
      `WG-Seq`,
      `WG-Nanopore`,
      harvested,
      species_in,
      selected,
      rep_type,
      rep,
      CT_total,
      n_passages,
      rep_type,
      rep,
      species_in,
      selected
    )
  ) %>%
  arrange(species, -selected)
```


```{r}

write_sheet(selected_samples_short, ss = "https://docs.google.com/spreadsheets/d/1csFMpLFtTr1_-hF1Nncduzjl9uIjKHRh0w7kntRZM2A/edit?usp=drive_link", 
            sheet = "selected_cell_lines")

write_sheet(selected_samples, ss = "https://docs.google.com/spreadsheets/d/1csFMpLFtTr1_-hF1Nncduzjl9uIjKHRh0w7kntRZM2A/edit?usp=drive_link",
            sheet = "selected_cell_lines_full")


```

```{r}

sum<-species_list %>% 
  select(species,individuals,superclass,genome_match,fibro_l,cat,project)

table(sum$project,sum$fibro_l)
table(sum$project,sum$superclass)

```



# Lines to harvest

```{r eval=FALSE, include=FALSE}
missing_species <- setdiff(subset(selection, silencer)$species, summary$species)

missing_species


new_lines <- summary %>%
  filter(lines_harvested < 2) %>%
  select(species, lines_tested) %>%
  left_join(selected_samples) %>%
  filter(!(animal_id %in% harvesting$animal_id)) %>%
  filter(!is.na(animal_id)) %>%
  
  write_sheet(
    "https://docs.google.com/spreadsheets/d/1vyBZwssG8ndUcGmDg4kLEPwOJJk15bOrXztO876HCOs/edit?usp=sharing",
    "new_lines"
  )

harvesting %>%
  group_by(animal_id, species) %>%
  summarize(tries = n(), Harvested = sum(Harvested) > 0) %>%
  filter(tries < 2 & !Harvested) %>%
  left_join(selected_samples) %>%
  filter(CT_total > 2) %>%
  
  
  write_sheet(
    "https://docs.google.com/spreadsheets/d/1vyBZwssG8ndUcGmDg4kLEPwOJJk15bOrXztO876HCOs/edit?usp=sharing",
    "repeat"
  )
```



```{r eval=FALSE, include=FALSE}
selected_samples2 <- subset(selected_samples, selected)

table(selected_samples2$sex)
table(selected_samples2$origin)

hist(table(selected_samples2$species))



selected_samples_full <- bind_rows(selected_samples2,
                                   selected_samples2[which(selected_samples2$reps == "techrep1"), ] %>%
                                     mutate(reps = "techrep2"))

hist(table(selected_samples_full$species))

```

# Plot 

```{r eval=FALSE, include=FALSE}


plot_df <- selected_samples %>%
  mutate(ispellet = pellet > 0) %>%
  select(species,
         animal_id,
         RNA_ex,
         from_cryo,
         silencer,
         illumina,
         pellet_RNA) %>%
  rownames_to_column(var = "index") %>%
  pivot_longer(
    cols = c(pellet_RNA, RNA_ex, from_cryo),
    names_to = "type",
    values_to = "value"
  ) %>%
  mutate(
    index = as.numeric(index),
    quarter = case_when(
      index < nrow(selected_samples) / 4 ~ "top",
      index < nrow(selected_samples) / 2 ~ "middle1",
      index > nrow(selected_samples) / 2 &
        index < (nrow(selected_samples) / 4) * 3 ~ "middle2",
      T ~ "bottom"
    ),
    type = factor(type, levels = rev(c(
      "from_cryo", "pellet_RNA", "RNA_ex"
    )))
  )

table(plot_df$quarter)


plot_df[is.na(plot_df)] <- F

ggplot(plot_df) +
  geom_tile(aes(
    x = type,
    y = reorder(animal_id, value),
    fill = value
  ), col = "white") +
  facet_wrap( ~ silencer, ncol = 4, scales = "free") +
  theme(
    strip.text = element_blank(),
    strip.background = element_blank(),
    axis.text.x = element_text(angle = 90, vjust = 0.3)
  ) +
  ggsci::scale_fill_npg() +
  xlab("")


table(plot_df$type, plot_df$value)
```

## Find Cryo tubes


```{r eval=FALSE, include=FALSE}

extr_long <- readRDS(paste0(path, "/data/Extractions_long.rds"))


IDs <- c(
  "A424
A495
A70
A460
A308
A489
A522
A37
A391
A301
A453
A357
A422
A65
A462
A530
A288
A461
A469
A345
A526
A583
A592
A441
A362
A185
A239
A172
A344
A143
A472
A270
A470
A394
A287
A302
A323
A204
A182
A361
A537
A317
A304
A452
A382
A132
A354
A411
A297
A48
A248"
) %>%
  str_split(pattern = "\n", simplify = F) %>%
  unlist()

positions <- extr_long %>%
  filter(type == "DNA") %>%
  mutate(column = as.numeric(str_extract(column, "\\d+"))) %>%
  mutate(row = as.numeric(str_extract(row, "\\d+"))) %>%
  mutate(animal_id = str_extract(cryotube, pattern = "[Aa]\\d+")) %>%
  filter(animal_id %in% IDs)

ggplot(positions) +
  geom_raster(aes(column, row)) +
  geom_hline(yintercept = 0.5:10.5) +
  geom_vline(xintercept = 0.5:10.5) +
  scale_y_reverse(breaks = 1:10) +
  scale_x_continuous(breaks = 1:12, position = "top") +
  geom_text(aes(column, row, label = cryotube),
            col = "white",
            size = 2) +
  facet_wrap( ~ box)



write_sheet(positions,"https://docs.google.com/spreadsheets/d/1FLSAxeA95XpqretA-YMJ22CQ6GWfBqF3zS96Iq_H2IM/edit?usp=sharing",sheet="Batch_1_DNA_locations")
```

