---
title: "1. Make Sample and Species Tables"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

In this script the tables from the Google sheet *List of Species* will be summarized into one species table and one sample table. The species table contains all the information on a species level such as, genome availability, iucn status, taxonomy and number and type of samples. The sample table will contain information about the individual samples such as sex, age, tissue of origin, storage location etc.

Those tables will be generated for mammals as well as for the remaining tetrapods (Birds,Reptiles, Amphibians).



```{r,message=F,warning=F}

library(tidyverse)
library(cowplot)
library(readxl)
library(runner)
library(janitor)
library(udpipe)
library(googlesheets4)

theme_set(theme_classic())

here::here()

path<-here::here()

```

This script generates a summary of the Cryozoo based on the "List of Species Sheet" 

This is the current status at : `r file.info(paste0(path,"/data/List of species.xlsx"))$ctime`

# Getting Species level table
## Load all species summary table
Import table:
```{r}

List_of_species <- read_excel(paste0(path,"/data/List of species.xlsx"), 
    sheet = "SPECIES LIST") %>% 
  distinct()

# rename ambiguous columns
colnames(List_of_species)[c(7,18)]<-c("individuals","tissue_of_origin")
colnames(List_of_species)<-janitor::make_clean_names(colnames(List_of_species))

### replace NA names
List_of_species[is.na(List_of_species$scientific_name),]$common_name
List_of_species$scientific_name[is.na(List_of_species$scientific_name)]<-List_of_species[is.na(List_of_species$scientific_name),]$common_name

```

## Tidy up table:

### 1. Replace problematic species names
```{r}
species_names<-read_excel(paste0(path,"/data/Wrong_Species_names.xlsx"),sheet = "names")

# Filter table to only include the entries that have a wrong name in the species table
species_names<-species_names[match(List_of_species$scientific_name[
  which(List_of_species$scientific_name%in%species_names$scientific_name_species)],
  species_names$scientific_name_species,
  incomparables = F),] %>% 
  mutate(scientific_name=scientific_name_species) %>% 
  distinct()

# Join with the list of species and replace with the new name, save the old name as alternative name
List_of_species<-List_of_species %>% 
  left_join(species_names) %>% 
  mutate(scientific_name=if_else(!(is.na(correct_name)),
                                   correct_name,
                                   scientific_name),
         alternative_name=if_else(scientific_name==correct_name,
                                   scientific_name_species,
                                   scientific_name)) %>%
  mutate(scientific_name=str_replace(scientific_name,pattern = "  ",replacement = " ")) %>% 
  dplyr::select(-c(correct_name,scientific_name_sample,scientific_name_species,group,wrong_sample,wrong_species,wrong_both))
```

## make naming reproducible
```{r}
List_of_species<-List_of_species %>% 
  rename(superclass=class) %>% 
  mutate(superclass=str_to_title(superclass),
         superclass=if_else(superclass=="Bird","Birds",superclass)) %>% 
  mutate(cell_line=str_replace(cell_line,pattern = "x",replacement = "Fibro")) %>% 
  mutate(cell_line=str_replace(cell_line,pattern = "X",replacement = "Fibro")) %>% 
  mutate(cell_line=str_replace(cell_line,pattern = "PBMc/Fibro",replacement = "Fibro/PBMc"))%>%
  mutate(iucn_category=if_else(iucn_category %in% c(NA,"NOT APPLICABLE","DD","??"),"NA",iucn_category))%>% 
  mutate(iucn_category=factor(iucn_category,levels=c("NA","LEAST CONCERN","VULNERABLE","NEAR THREATENED","ENDANGERED","CRITICALLY ENDANGERED","EXTINCT IN THE WILD"))) %>% 
  mutate(individuals=as.numeric(individuals),
         individuals=if_else(is.na(individuals),1,individuals)) %>% 
  mutate(order=as.factor(order)) %>% 
  mutate(scientific_name=str_replace(scientific_name,pattern = "\\?",replacement = ""),
         scientific_name=str_to_sentence(scientific_name)) %>%  
  separate(scientific_name,into=c("n1","n2","subspecies"),remove = F) %>% 
  mutate(n1=stringr::str_to_title(n1),
         n2=tolower(n2)) %>% 
  unite(c(n1,n2),col = "species",sep=" ") %>% 
  arrange(order) 

```



## Check for duplicated names
```{r}

bind_rows(List_of_species[duplicated(List_of_species$species),],
          List_of_species[duplicated(List_of_species$species,fromLast = T),])

```

Some of those are truly duplicate entries, some are subspecies that for our purpose will be merged

## remove duplicate entries for species

```{r}
# get species names of duplicated entries
dup_species<-List_of_species[duplicated(List_of_species$species),]$species

# filter list of species for those and collapse the information to not loose anything, finally remove the duplicates
dup_sum <-
  List_of_species %>% 
  filter(species%in% dup_species
  ) %>%
  group_by(species) %>% 
  summarize(
    subspecies = paste(
      do::unique_no.NA(subspecies),collapse = ", "
      ),
    subfamily = paste(
      do::unique_no.NA(subfamily), collapse= ", "
      ),
    cites_annex = paste(
      do::unique_no.NA(cites_annex), collapse= ", "
      ),
    sex = paste(
      do::unique_no.NA(sex), collapse= ", "
      ),
    superclass = paste(
      do::unique_no.NA(superclass), collapse= ", "
      ),
    alternative_name = paste(
      do::unique_no.NA(alternative_name), collapse= ", "
      ),
    common_name = paste(
      do::unique_no.NA(common_name), collapse= ", "
      ),
    cell_line = paste(
      do::unique_no.NA(cell_line), collapse= ", "
      ),
    tissue_of_origin = paste(
      do::unique_no.NA(tissue_of_origin), collapse= ", "
      ),
    iucn_category =  paste(
      do::unique_no.NA(as.character(iucn_category)), collapse= ", "
      ),
    karyotype =  paste(
      do::unique_no.NA(karyotype), collapse= ", "
      ),
    dna =  paste(
      do::unique_no.NA(dna), collapse= ", "
      ),
    rna = paste(
      do::unique_no.NA(rna), collapse= ", "
      ),
    wgs_sequencing =  paste(
      do::unique_no.NA(wgs_sequencing), collapse= ", "
      ),
    rna_seq =  paste(
      do::unique_no.NA(rna_seq), collapse= ", "
      ),
    i_psc =  paste(
      do::unique_no.NA(i_psc), collapse= ", "
      ),
    order = paste(
      do::unique_no.NA(as.character(order)), collapse= ", "
      ),
    individuals = sum(individuals),
    scientific_name = species
  ) %>% 
  filter(duplicated(species))


# add this back to the main list
List_of_species<-List_of_species %>% 
  filter(!(species %in% dup_species)) %>% 
  bind_rows(dup_sum) %>% 
  mutate(order=as.character(order)) %>% 
  mutate(iucn_category=if_else(iucn_category=="ENDANGERED, NA","ENDANGERED",iucn_category)) %>% 
  mutate(species_=str_replace(species,pattern = " ",replacement = "_")) %>% 
  mutate(cell_line=if_else(is.na(cell_line),"No cell line",cell_line))

```



```{r}
head(List_of_species)

saveRDS(List_of_species,paste0(path,"/output/Species_List.rds"))
```


## Add genome information from Genome on a tree

```{r}
## load list genomes from Goat
info <- file.info(paste0(path,"/data/GenomeOnATree_Chordata.tsv"))
mod_time <- info$mtime
print(paste0("downloaded from GOAT ",mod_time))
GOAT_info <- read_delim(paste0(path,"/data/GenomeOnATree_Chordata.tsv"),
  delim = "\t", escape_double = FALSE,
  col_types = cols(taxon_id = col_character()),
  trim_ws = TRUE) %>%
  mutate(common_name=str_replace_all(common_name, c('\\[\"'="",'\"\\]'=""))) %>%
  mutate(synonym=str_replace_all(synonym, c('\\[\"'="",'\"\\]'=""))) %>%
  mutate(tolid_prefix=str_replace_all(tolid_prefix, c('\\[\"'="",'\"\\]'=""))) %>%
  mutate(assembly_level=str_to_title(assembly_level))
# %>%
# group_by(species) %>%
# mutate(best_asm=scaffold_n50==max(scaffold_n50,na.rm=T)) %>%
# mutate(best_asm=if_else(is.na(best_asm),T,best_asm)) %>%
# ungroup() %>%
# filter(best_asm) %>%
# select(-scientific_name,-best_asm) %>%
# distinct()

```


```{r}
# annotation info
# this is pretty outdated
info <- file.info(paste0(path,"/data/Ensembl_NCBI_annotations.rds"))
mod_time <- info$mtime
print(paste0("downloaded from Ensembl/NCBI ",mod_time))

anno<-readRDS(paste0(path,"/data/Ensembl_NCBI_annotations.rds")) %>% 
  filter(!duplicated(Species))
```


## add TOGA annotation
```{r}
Toga1<-read_sheet("https://docs.google.com/spreadsheets/d/1woHdlYmgVm_QoVDb5tokMboENvOTTF8fuGcoSOIJz0A/edit?usp=sharing",sheet = "Hiller_lab_TOGA1") %>%
janitor::clean_names() %>%
select(species,species_taxonomy_id,assembly_name,contig_n50_bp,scaffold_n50_bp) %>%
mutate(toga1=T) %>%
mutate(species_taxonomy_id=as.numeric(species_taxonomy_id))

Toga2<-read_sheet("https://docs.google.com/spreadsheets/d/1woHdlYmgVm_QoVDb5tokMboENvOTTF8fuGcoSOIJz0A/edit?usp=sharing",sheet = "Hiller_lab_TOGA2") %>%
janitor::clean_names() %>%
select(species,species_taxonomy_id,assembly_name,contig_n50_bp,scaffold_n50_bp) %>%
mutate(toga2=T)

TOGA_ann<-full_join(Toga2,Toga1) %>%
  mutate(species=str_remove(species, "_[^_]+$")) %>% 
group_by(species) %>%
slice_max(scaffold_n50_bp) %>%
slice_max(contig_n50_bp) %>%
mutate(toga1=sum(toga1,na.rm = T)>0) %>%
mutate(toga2=sum(toga2,na.rm = T)>0) %>%
select(-assembly_name) %>%
distinct()

table(TOGA_ann$toga1,TOGA_ann$toga2,useNA = "always")

TOGA_ann[duplicated(TOGA_ann$species),]

```


```{r}
## join with genome info
GOAT_info<-left_join(GOAT_info,anno,by=c("scientific_name"="Species")) %>%
full_join(select(TOGA_ann,c(species,toga1,toga2)))

# check if any species from the cryozoo is not in the GOAT database
table((List_of_species$scientific_name %in% GOAT_info$scientific_name), List_of_species$superclass,useNA = "always")

# if so, which?
List_of_species[!(List_of_species$species %in% GOAT_info$species),]

List_of_species_new<-left_join(select(List_of_species,-c(family,order)),select(GOAT_info,-scientific_name),by=c("species")) %>% 
  mutate(common_name=case_when(is.na(common_name.x)~common_name.y,
                               common_name.x==common_name.y~common_name.y,
                               common_name.x!=common_name.y~paste(common_name.x,common_name.y,sep=","),
                               T~NA_character_)#,
         # family=case_when(is.na(family.y)~family.x,
         #                  is.na(family.x)~family.y,
         #                       family.x==family.y~family.y,
         #                       family.x!=family.y~family.y,
         #                       T~NA_character_),
         # order=case_when(is.na(order.y)~order.x,
         #                 is.na(order.x)~order.y,
         #                       order.x==order.y~order.y,
         #                       order.x!=order.y~order.y,
         #                       T~NA_character_)
         ) %>% 
  select(-c(ends_with(".x"),ends_with(".y"),taxon_rank)) %>% 
  mutate(taxon_id=if_else(is.na(taxon_id),"not in genome list",taxon_id)) 


saveRDS(List_of_species_new,paste0(path,"/output/Species_List_with_genome_info.rds"))
```


```{r}
## match species that do not yet have a genome assembly to their closest genome

# keep those columns that refer to the genome rather than the reference
genome_columns<-c("species","genus","family","superclass","taxon_id","tolid_prefix","c_value","genome_size","chromosome_number","haploid_number","ploidy")

# keep those columns that refer to the assembly not the reference
assembly_columns<-c("taxon_id","assembly_id","assembly_name","genbank_accession","refseq_accession",
                    "species","genus","family","synonym","common_name","order","class",
                    "phylum","kingdom","superkingdom","genome_size_draft","contig_n50",
                    "gene_count","assembly_date","assembly_level","assembly_span","scaffold_n50",
                    "busco_completeness","published","sequencing_status","Name","Annotation.Release",
                    "Freeze.Date","Release.Date","Common.name","Taxon.ID","Ensembl.Assembly","Accession",
                    "Genebuild.Method","Variation.database","Regulation.database","toga1","toga2")

## join the species with is.na(assembly_level) with GOAT by genus and select the genome of the same genus with the highest N50
alt_genomes<-List_of_species_new %>% 
  filter(is.na(assembly_level)) %>% 
  select(all_of(genome_columns))%>% 
  rename(taxon_id_sp=taxon_id) %>% 
  left_join(subset(select(GOAT_info,assembly_columns),!is.na(assembly_level)),by="genus") %>% 
  rename(taxon_id_ref=taxon_id,
         taxon_id=taxon_id_sp) %>% 
  group_by(species.x) %>% 
  slice_max(scaffold_n50) %>% 
  mutate(genome_match=if_else(is.na(assembly_level),NA_character_,"genus")) %>% 
  rename(species_genome=species.y,
         family=family.x,
         species=species.x) %>% 
  select(-family.y)

## join the species with is.na(assembly_level) with GOAT by family and select the genome of the same family with the highest N50
alt_genomes2<-alt_genomes %>% 
  filter(is.na(assembly_level)) %>% 
  select(all_of(genome_columns)) %>% 
  rename(taxon_id_sp=taxon_id) %>% 
  left_join(subset(select(GOAT_info,assembly_columns,-genus),!is.na(assembly_level)),by="family") %>% 
  rename(taxon_id_ref=taxon_id,
         taxon_id=taxon_id_sp) %>% 
  group_by(species.x) %>% 
  slice_max(scaffold_n50) %>% 
  mutate(genome_match=if_else(is.na(assembly_level),NA_character_,"family"))  %>% 
  rename(species_genome=species.y,
         species=species.x) 

# combine genus level and family level matches
alt_genomes_full<-rbind(subset(alt_genomes,!is.na(assembly_level)),subset(alt_genomes2,!is.na(assembly_level))) %>% 
  select(-any_of(c("synonym", "genus", "phylum", "kingdom", "superkingdom", "common_name")))

genome_colnames<-c("taxon_rank","tolid_prefix","c_value","genome_size","genome_size_draft","contig_n50","gene_count","assembly_id","assembly_name","genbank_accession","refseq_accession","assembly_date","assembly_level","assembly_span","scaffold_n50","busco_completeness","chromosome_number","haploid_number","ploidy","published","sequencing_status","Name","Annotation.Release","Freeze.Date","Release.Date","Common.name","Taxon.ID","Ensembl.Assembly","Accession","Genebuild.Method","Variation.database","Regulation.database","toga1","toga2")

## for these species remove the reference genome info and replace it with the closest ref genome
List_of_species_alt_genomes<-subset(List_of_species_new,species%in%alt_genomes_full$species) %>% 
  select(-any_of(which(colnames(List_of_species_new)%in%genome_colnames))) %>% 
  left_join(alt_genomes_full)

## add the remaining sepcies with ref genomes back to the list 
List_of_species_alt_genomes<-subset(List_of_species_new,!(species%in%alt_genomes_full$species)) %>% 
  mutate(genome_match=if_else(is.na(assembly_level),"none","species"),
         species_genome=if_else(is.na(assembly_level),"none",species)) %>% 
  bind_rows(List_of_species_alt_genomes)


table(List_of_species_alt_genomes$genome_match,List_of_species_alt_genomes$superclass,useNA = "always")
```


## Add missing divergence times from time tree

```{r}

## already collected previously
divergence_time<-read_delim(paste0(path,"/data/species_time_tree.csv"),
                            delim = ",",
                            skip_empty_rows = T)[,1:6]

query_timetree<-List_of_species_alt_genomes %>%
  filter(!(genome_match%in%c("species","none"))) %>% 
  select(species,taxon_id,species_genome,taxon_id_ref,genome_match)


```

## get divergence times using the timetree API
```{r}

library(httr)
library(purrr)

get_divergence_time <- function(taxon_id, taxon_id_ref) {
  #taxon_id<-query_timetree$taxon_id
  #taxon_id_ref<-query_timetree$taxon_id_ref
  url <- paste0("http://timetree.temple.edu/api/pairwise/", taxon_id, "/", taxon_id_ref)
  res <- GET(url)
  stop_for_status(res)
  
  txt <- content(res, as = "text", encoding = "UTF-8")
  
  if (grepl("Undefined index", txt) || grepl("<br", txt, fixed = TRUE) || txt=="") {
    return(tibble(
      taxon_a_id = taxon_id,
      taxon_b_id = taxon_id_ref,
      status = "No match"
    ))
  }
  
  df <- read_csv(txt, show_col_types = FALSE) %>%
    mutate(status = "OK") %>% 
    mutate(across(starts_with("taxon_"),as.character))
  return(df)
}



```


## divergence time from time tree
```{r}
# Apply directly to dataframe
results <- pmap_dfr(select(query_timetree,c(taxon_id,taxon_id_ref)), get_divergence_time)

table(results$status)

results<-select(results,c("taxon_a_id","taxon_b_id","status","precomputed_age","adjusted_age")) %>% 
  rename(taxon_id=taxon_a_id,
         taxon_id_ref=taxon_b_id)

divergence_time<- query_timetree %>% 
  left_join(results) %>% 
  mutate(divergence_mya_adj=if_else(is.na(precomputed_age),
                                    if_else(genome_match=="genus",10,20),
                                    precomputed_age)) %>%  
  inner_join(alt_genomes_full) %>% 
  mutate()

divergence_time%>% 
  ggplot()+
  geom_density(aes(x=divergence_mya_adj,col=genome_match))+
  scale_x_log10()+
  geom_vline(xintercept = 10,lty="dashed")+
  facet_wrap(~superclass)

table(divergence_time$divergence_mya_adj<10,divergence_time$genome_match)
```


# add genome info to main table for divergence times < 10 mya

```{r}

List_of_species_alt_genomes2<-subset(List_of_species_new,species%in%alt_genomes_full$species) %>%
  select(-which(colnames(List_of_species_new)%in%genome_colnames)) %>%
  left_join(divergence_time)

List_of_species_alt_genomes2<-subset(List_of_species_new,!(species%in%alt_genomes_full$species)) %>% 
  mutate(genome_match=if_else(is.na(assembly_level),"none","species"),
         species_genome=if_else(is.na(assembly_level),"none",species),
         divergence_mya_adj=0.0001) %>% 
  bind_rows(List_of_species_alt_genomes2)

 ggplot()+
  geom_histogram(data=List_of_species_alt_genomes2,aes(x=divergence_mya_adj,fill=genome_match))+
  geom_vline(xintercept = 10,lty="dashed")+
  facet_wrap(~superclass)

 
```


### add zoonomia info

```{r}

Zoonomia_species_Table_S1 <- read_excel(paste0(path,"/data/Zoonomia_species_Table_S1.xlsx")) %>% 
  separate(Species,into=c("n1","n2","subspecies"),remove = F) %>% 
  mutate(n1=stringr::str_to_title(n1),
         n2=tolower(n2)) %>% 
  unite(c(n1,n2),col = "scientific_name",sep=" ")
  
### some names have to be changed to be matched to GOAT

Zoonomia_species_Table_S1 <- read_excel(paste0(path,"/data/Zoonomia_goat_names.xlsx")) %>% 
  right_join(Zoonomia_species_Table_S1) %>% 
  mutate(species=if_else(is.na(goat_name),scientific_name,goat_name)) %>% 
  select(-goat_name)

List_of_species_new_zoo<-List_of_species_alt_genomes2 %>%
  mutate(zoonomia=species %in% Zoonomia_species_Table_S1$species)%>% 
  mutate(genome=divergence_mya_adj<10,
         genome=if_else(genome,genome,zoonomia),
         annotation=zoonomia==T| !is.na(Annotation.Release)| !is.na(Ensembl.Assembly)| !is.na(gene_count)) 


table(List_of_species_new_zoo$zoonomia,List_of_species_new_zoo$sequencing_status)

```


# representation of cryzoo species

```{r}

z<-Zoonomia_species_Table_S1%>% 
  select(scientific_name) %>% 
  rename(species=scientific_name) %>% 
  left_join(GOAT_info) %>% 
  select(species,genus,family,order,class,phylum) %>% 
  filter(!is.na(order)) %>% 
  mutate(group="zoonomia",
         zoonomia=T) 


a<-GOAT_info%>% 
  filter(!is.na(order)) %>% 
  select(species,genus,family,order,class,phylum) %>% 
  mutate(group="all_mammals",
         all_mammals=T)

c<-List_of_species %>% 
  select(species,cell_line) %>% 
  left_join(GOAT_info[,c("species","genus","family","order","class","phylum")]) %>% 
  select(species,genus,family,order,class,phylum,cell_line) %>% 
  mutate(group="cryozoo",
         cryozoo=T) %>% 
  select(-cell_line)

all_species_summary<-rbind(z[,1:7],a[,1:7],c[,1:7]) %>% 
  filter(class=="Mammalia") %>% 
  group_by(group,class,phylum,order) %>% 
  summarize(nspecies=length(unique(species)),ngenera=length(unique(genus)),nfamily=length(unique(family)))
```


```{r}
ggplot(all_species_summary)+
  geom_col(aes(y=nspecies,x=group),position = position_dodge())+
  facet_wrap(~order,scales = "free_x")+
  coord_flip()
```


```{r}
ggplot(all_species_summary)+
  geom_col(aes(y=ngenera,x=group),position = position_dodge())+
  facet_wrap(~order,scales = "free_x")+
  coord_flip()
```


```{r}
ggplot(all_species_summary)+
  geom_col(aes(y=nfamily,x=group),position = position_dodge())+
  facet_wrap(~order)+
  coord_flip()
```


```{r}
all_mammals<-all_species_summary %>% 
  filter(group=="all_mammals") %>% 
  ungroup() %>% 
  select(-c(group)) %>% 
  rename(total_genera=ngenera,
         total_species=nspecies,
         total_families=nfamily)
  
 all_species_summary<-all_species_summary %>% 
   filter(group!="all_mammals") %>% 
   left_join(all_mammals) %>% 
   mutate(p_species=(nspecies/total_species)*100,
          p_genera=(ngenera/total_genera)*100,
          p_families=(nfamily/total_families)*100)
    
  ggplot()+
    ggbeeswarm::geom_beeswarm(data=all_species_summary,aes(x=reorder(order,total_species),col=group,y=p_species,size=total_species))+
    coord_flip()+
    theme_light()+
      scale_colour_manual(values=c("zoonomia"="#e63946","cryozoo"="#6987b5"))+
      labs(x="mammalian orders\n (sorted by number of species)",y="percent of all species covered")
```


```{r}
sum_long<-all_species_summary %>% 
  pivot_longer(cols=c(p_species,p_genera,p_families),names_to = c("P","tax_level"),values_to = "percent",names_sep = "_") %>% 
  select(-P) %>% 
  mutate(tax_level=factor(tax_level,levels=rev(c("families","genera","species"))))

ggplot()+
    ggbeeswarm::geom_beeswarm(data=sum_long,aes(x=reorder(order,total_species),colour=group,y=percent),position=position_dodge())+
    coord_flip()+
    theme_light()+
      scale_colour_manual(values=c("zoonomia"="#e63946","cryozoo"="#6987b5"))+
      labs(x="mammalian orders\n (sorted by number of species)",y="percent of all taxonomic units covered") +
  facet_grid(~tax_level)
  
```

# Summary plots genome

## Sequencing Status
```{r}

# per superclass

List_of_species_new_zoo %>% 
  mutate(sequencing_status=factor(sequencing_status,levels=rev(c("sample_collected","sample_acquired","in_progress","open","insdc_open","published")))) %>% 
  mutate(superclass=factor(superclass,levels=rev(c("Mammalia","Birds","Amphibia","Reptil","Fish")))) %>% 

ggplot()+
  geom_bar(aes(fill=sequencing_status,x=superclass))+
  coord_flip()

# per order

List_of_species_new_zoo %>% 
  mutate(sequencing_status=factor(sequencing_status,levels=rev(c("sample_collected","sample_acquired","in_progress","open","insdc_open","published")))) %>% 
   mutate(superclass=factor(superclass,levels=rev(c("Mammalia","Birds","Amphibia","Reptil","Fish")))) %>% 
ggplot()+
  geom_bar(aes(fill=sequencing_status,x=order))+
  coord_flip()+
  facet_grid(superclass~genome,scales = "free",space = "free") 



```

# Getting Sample level tables
## Sample table Mammals

```{r}

mammals_list <- read_excel(paste0(path,"/data/List of species.xlsx"), 
    sheet = "MAMMALIA",skip = 1)[-1,-2]

colnames(mammals_list)<-janitor::make_clean_names(colnames(mammals_list))

colnames(mammals_list)[2]<-"scientific_name"

##Make a clean tidy table for mammals
 

mammals_list_clean<-mammals_list %>% 
  mutate(x1=fill_run(x1,run_for_first = F),
         scientific_name=fill_run(scientific_name),
         common_name=fill_run(common_name),
         family=fill_run(family),
         iucn=fill_run(iucn)) %>% 
  filter(!(is.na(zims_species360)&is.na(sample_id))) %>% 
  mutate(sample_id=if_else(is.na(sample_id)& grepl(x = genetic_material,pattern = "^A"),
                           genetic_material,sample_id)) %>%  
  mutate(genetic_material=if_else(grepl(x = genetic_material,pattern = "^A"),
                                  NA_character_,genetic_material)) %>%  
  mutate(iPSC=grepl(genetic_material,pattern="iPSC")) %>% 
  mutate(sample_id= str_replace(sample_id,pattern="\\(",replacement = "/")) %>% 
  mutate(sample_id= str_replace(sample_id,pattern="\\)",replacement = "")) %>% 
  mutate(sample_id= str_replace(sample_id,pattern="\\s",replacement = "/")) %>% 
  separate(sample_id,sep = "/",into = c("line1","line2","line3"),remove = F) %>% 
  separate(genetic_material,sep = "/",into = c("cell_type","type1","type2","type3","type4")) %>% 
  mutate(cell_type=case_when(cell_type=="C"~"Fibro",
                             cell_type=="PBMC"~"PBMc",
                             cell_type=="BLOOD"~"PBMc",
                             cell_type=="PMBc isolation"~"PBMc",
                             cell_type=="?"~"unclear",
                             cell_type=="-"~"unclear",
                             is.na(cell_type)~"Primary",
                             T~"Primary")) %>% 
  mutate(status=cell_type!="Primary") %>% 
  mutate(scientific_name=str_replace(scientific_name,pattern = "\\?",replacement = ""),
         scientific_name=str_to_sentence(scientific_name),
         superclass="Mammalia")
  

# correct bad names
species_names_mammals<-subset(species_names,group=="Mammals") %>% 
  mutate(scientific_name_sample=str_to_sentence(scientific_name_sample))

species_names_mammals<-species_names_mammals[match(mammals_list_clean$scientific_name[
  which(mammals_list_clean$scientific_name%in%species_names_mammals$scientific_name_sample)],
  species_names_mammals$scientific_name_sample,
  incomparables = F),]

mammals_list_clean$scientific_name[which(
  mammals_list_clean$scientific_name%in%species_names_mammals$scientific_name_sample)
  ]<-species_names_mammals$correct_name


mammals_list_clean<-mammals_list_clean %>% 
  separate(scientific_name,into=c("n1","n2","subspecies"),remove = F) %>% 
  mutate(n1=stringr::str_to_title(n1),
         n2=tolower(n2)) %>% 
  unite(c(n1,n2),col = "species",sep=" ")

saveRDS(mammals_list_clean,paste0(path,"/output/Sample_List_Mammal_clean.rds"))
write_excel_csv(mammals_list_clean,paste0(path,"/output/Sample_List_Mammal_clean.csv"))


```


## Sample Table other Tetrapods

```{r}

tetra_list <- read_excel(paste0(path,"/data/List of species.xlsx"), 
    sheet = "AMPHIBIAN",
    col_names = T,skip = 1)[-1,-2]

tetra_list$superclass<-"Amphibian"

tetra_list<-bind_rows(tetra_list,
                      read_excel(paste0(path,"/data/List of species.xlsx"), 
    sheet = "REPTIL",col_names = T,skip = 1)[-c(1),-2])

tetra_list$superclass[is.na(tetra_list$superclass)]<-"Reptiles"

bird_list<- read_excel(paste0(path,"/data/List of species.xlsx"), 
    sheet = "BIRD",col_names = T,skip = 1)[-1,-2] %>% 
   rename('CELL BARCODE'='CELL LINE BARCODE')

tetra_list<-bind_rows(tetra_list,bird_list)


tetra_list$superclass[is.na(tetra_list$superclass)]<-"Birds"

colnames(tetra_list)<-janitor::make_clean_names(colnames(tetra_list))

colnames(tetra_list)[2]<-"scientific_name"

```

```{r}

tetra_list_clean<-tetra_list %>% 
  mutate(x1=fill_run(x1,run_for_first = F),
         scientific_name=fill_run(scientific_name),
         common_name=fill_run(common_name),
         family=fill_run(family),
         iucn=fill_run(iucn)) %>% 
  filter(!(is.na(zims_species360)&is.na(sample_id)&is.na(genetic_material))) %>% 
  mutate(sample_id= str_replace(sample_id,pattern="\\(",replacement = "/")) %>% 
  mutate(sample_id= str_replace(sample_id,pattern="\\)",replacement = "")) %>% 
  mutate(sample_id= str_replace(sample_id,pattern="\\s",replacement = "/")) %>% 
  separate(sample_id,sep = "/",into = c("line1","line2","line3"),remove = F) %>% 
  separate(genetic_material,sep = "/",into = c("cell_type","type1","type2","type3","type4")) %>% 
  mutate(cell_type=case_when(cell_type=="C"~"Fibro",
                             cell_type=="PBMC"~"PBMc",
                             cell_type=="BLOOD"~"PBMc",
                             cell_type=="PMBc isolation"~"PBMc",
                             cell_type=="?"~"unclear",
                             cell_type=="-"~"unclear",
                             is.na(cell_type)~"Primary",
                             T~"Primary")) %>% 
  mutate(status=cell_type!="Primary") %>% 
  mutate(scientific_name=str_replace(scientific_name,pattern = "\\?",replacement = ""),
         scientific_name=str_to_sentence(scientific_name)) %>% 
  mutate(iPSC=F)

### correct names
species_names_tetra<-subset(species_names,group!="Mammals") %>% 
  mutate(scientific_name_sample=str_to_sentence(scientific_name_sample))

species_names_tetra<-species_names_tetra[match(tetra_list_clean$scientific_name[
  which(tetra_list_clean$scientific_name%in%species_names_tetra$scientific_name_sample)],
  species_names_tetra$scientific_name_sample,
  incomparables = F),]

tetra_list_clean$scientific_name[which(
  tetra_list_clean$scientific_name%in%species_names_tetra$scientific_name_sample)
  ]<-species_names_tetra$correct_name

tetra_list_clean<-tetra_list_clean %>% 
  separate(scientific_name,into=c("n1","n2","subspecies"),remove = F) %>% 
  mutate(n1=stringr::str_to_title(n1),
         n2=tolower(n2)) %>% 
  unite(c(n1,n2),col = "species",sep=" ") %>% 
  filter(!duplicated(sample_id)) %>% 
  distinct()

saveRDS(tetra_list_clean,paste0(path,"/output/Sample_List_Tetrapod_clean.rds"))
write_excel_csv(tetra_list_clean,paste0(path,"/output/Sample_List_Tetrapod_clean.csv"))

```

## Sample Table Fish


```{r}

fish_list <- read_excel(paste0(path,"/data/List of species.xlsx"), 
    sheet = "FISH",
    col_names = T,skip=1)[-1,-2]

fish_list$superclass<-"Fish"

colnames(fish_list)<-janitor::make_clean_names(colnames(fish_list))
colnames(fish_list)[2]<-"scientific_name"
```

```{r}


fish_list_clean<-fish_list %>% 
  mutate(x1=fill_run(x1,run_for_first = F),
         scientific_name=fill_run(scientific_name),
         common_name=fill_run(common_name),
         family=fill_run(family),
         iucn=fill_run(iucn)) %>% 
  filter(!(is.na(zims_species360)&is.na(sample_id)&is.na(genetic_material))) %>% 
  mutate(sample_id= str_replace(sample_id,pattern="\\(",replacement = "/")) %>% 
  mutate(sample_id= str_replace(sample_id,pattern="\\)",replacement = "")) %>% 
  mutate(sample_id= str_replace(sample_id,pattern="\\s",replacement = "/")) %>% 
  separate(sample_id,sep = "/",into = c("line1","line2","line3"),remove = F) %>% 
  separate(genetic_material,sep = "/",into = c("cell_type","type1","type2","type3","type4")) %>% 
  mutate(cell_type=case_when(cell_type=="C"~"Fibro",
                             cell_type=="PBMC"~"PBMc",
                             cell_type=="BLOOD"~"PBMc",
                             cell_type=="PMBc isolation"~"PBMc",
                             cell_type=="?"~"unclear",
                             cell_type=="-"~"unclear",
                             is.na(cell_type)~"Primary",
                             T~"Primary")) %>% 
  mutate(status=cell_type!="Primary") %>% 
  mutate(scientific_name=str_replace(scientific_name,pattern = "\\?",replacement = ""),
         scientific_name=str_to_sentence(scientific_name)) %>% 
  mutate(iPSC=F)

### correct names
species_names_fish<-subset(species_names,group=="Fish") %>% 
  mutate(scientific_name_sample=str_to_sentence(scientific_name_sample))

species_names_fish<-species_names_fish[match(fish_list_clean$scientific_name[
  which(fish_list_clean$scientific_name%in%species_names_fish$scientific_name_sample)],
  species_names_fish$scientific_name_sample,
  incomparables = F),]

fish_list_clean$scientific_name[which(
  fish_list_clean$scientific_name%in%species_names_fish$scientific_name_sample)
  ]<-species_names_fish$correct_name

fish_list_clean<-fish_list_clean %>% 
  separate(scientific_name,into=c("n1","n2","subspecies"),remove = F) %>% 
  mutate(n1=stringr::str_to_title(n1),
         n2=tolower(n2)) %>% 
  unite(c(n1,n2),col = "species",sep=" ")

saveRDS(fish_list_clean,paste0(path,"/output/Sample_List_fish_clean.rds"))
write_excel_csv(fish_list_clean,paste0(path,"/output/Sample_List_fish_clean.csv"))

```


## Combine sample tables

```{r}
combined_sample<-bind_rows(tetra_list_clean,mammals_list_clean) %>% 
  bind_rows(fish_list_clean) %>% 
  select(-c(x1,cell_barcode)) %>% 
  mutate(species_short=paste(str_to_title(substr(str_split(string = species,
                                                           pattern = " ",
                                                           simplify=T)[,1],1,3)),
                             str_to_title(substr(str_split(string = species,
                                                           pattern = " ",
                                                           simplify=T)[,2],1,3)),sep="_")) %>% 
  group_by(species_short) %>% 
  mutate(sample_id=paste(sep="_",species_short,row_number())) %>% 
  rename(animal_ID=line1) %>% 
  ungroup() %>% 
  mutate(link=txt_paste(x14,x15,data,na.rm = T)) %>% 
  select(-c(x14,x15,data)) %>% 
  mutate(animal_ID=str_remove(animal_ID,pattern="\\?"),
         dna=if_any(.cols=starts_with("type"),~grepl("DNA",.)),
         karyo=if_any(.cols=starts_with("type"),~grepl("K",.)),
         rna=if_any(.cols=starts_with("type"),~grepl("RNA",.)),
         iPSC2=if_any(.cols=starts_with("type"),~grepl("iPSC",.)),
         seq=if_any(.cols=starts_with("type"),~grepl("SEQ",.))) %>% 
  mutate(iPSC=if_else(iPSC==iPSC2,iPSC,T)) %>% 
  select(-c(iPSC2,type1,type2,type3,type4))
  
  
saveRDS(combined_sample,
       paste0(path,"/output/Sample_List_all_species_clean.rds"))
write_excel_csv(combined_sample,
                paste0(path,"/output/Sample_List_all_species_clean.csv"))

```

Add Info about cell lines to Species Table:

```{r}
summary_species<-combined_sample %>% 
  group_by(species) %>% 
  summarize(individuals_cl=n(),
         cell_lines=sum(status),
         fibro=sum(cell_type=="Fibro"),
         pbmc=sum(cell_type=="PBMc"),
         ipsc=sum(iPSC),
         scientific_name=paste(unique(scientific_name), collapse = ", "),
         subspecies=paste(unique(subspecies), collapse = ", "),
         common_name=paste(unique(common_name), collapse = ", "),
         family=paste(unique(family), collapse = ", ")) %>% 
ungroup()

Species_list_complete<-List_of_species_new_zoo%>% 
  full_join(select(summary_species,c(species,individuals_cl,cell_lines,fibro,pbmc,ipsc))) %>% 
  mutate(fibro=if_else(is.na(fibro),0,fibro),
         cell_lines=if_else(is.na(cell_lines),0,cell_lines),
         pbmc=if_else(is.na(pbmc),0,pbmc),
         ipsc=if_else(is.na(ipsc),0,ipsc)
         )

saveRDS(Species_list_complete,paste0(path,"/output/Species_List_with_genome_and_cl_info.rds"))

Species_list_complete %>% 
  select(taxon_id, species,	order,	class, superclass,	iucn_category,genome,	annotation, zoonomia,	individuals,cell_lines,fibro,pbmc,ipsc, 	tissue_of_origin, genome_match, species_genome,	assembly_level,	scaffold_n50,	contig_n50,	gene_count,	assembly_date,	busco_completeness,	genome_size,	haploid_number,	ploidy,	sequencing_status,	Annotation.Release,	Ensembl.Assembly,Accession, common_name,	alternative_name, subspecies,	genus,	subfamily,	family) %>% 
  
  write_excel_csv(paste0(path,"/output/List_of_species_all_info_short.csv"))

knitr::kable(table(Species_list_complete$fibro>0,Species_list_complete$genome))

```


## Compare sample list and species list info

Entries only found in species table: `r setdiff(List_of_species_new_zoo$species,summary_species$species)`

Entries only found in sample table:`r setdiff(summary_species$species,List_of_species_new_zoo$species)`


```{r}
List_of_species_new_zoo %>% 
  select(species,individuals,tissue_of_origin,sex,cell_line,i_psc) %>% 
  inner_join(summary_species) %>% 
  ggplot()+
  geom_point(aes(individuals,individuals_cl))+
  geom_abline(slope = 1,intercept = 0)

sum<-List_of_species %>% 
  select(species,individuals,tissue_of_origin,sex,cell_line,i_psc) %>% 
  inner_join(summary_species) %>% 
  mutate(comp=case_when(individuals==individuals_cl~"same",
                        individuals>individuals_cl~"more in species list",
                        individuals<individuals_cl~"more in sample list"
                        ))

table(sum$comp)

```

```{r}
work_table<-read_sheet("https://docs.google.com/spreadsheets/d/1SRid9RGRfcCpGAWsAXG8h0wE-7p1fhOtwvkMJgR1s8w/edit?usp=sharing",sheet = "Samples from A274",skip = 2,col_types = "ccccccccccccccccc") %>% 
  select(-c(3,13,16,17))

work_table<-work_table %>% 
  clean_names() %>% 
  filter(!if_all(everything(), is.na)) %>% 
  mutate(animal_species=if_else(animal_species%in%c("second_round","second round"),
                                NA_character_,
                                animal_species)) %>% 
  filter(date!="DATE") %>% 
  fill(c(samples,date,animal_species,coming_from,species_360_local_id),.direction = "down") %>% 
  distinct() %>% 
  mutate(coming_from_clean = case_when(
    # Acuario Gijon variations
    coming_from %in% c("Acuario de Gijon", "Acuario Gijon") ~ "Aquarium de Gijon",

    # Acuario Zaragoza
    coming_from == "Acuario de Zaragoza" ~ "Aquarium de Zaragoza",

    # BCN Zoo variations
    coming_from %in% c("BCN zoo", "BCN Zoo", "BCN ZOO", "BCN ZOO (frozen)", 
                       "BCN ZOO, Zoo Barcelona", "zoo Barcelona", 
                       "Zoo barcelona", "Zoo Barcelona", "Zoo de Barcelona","Potamochoerus porcus") ~ "Zoo Barcelona",

    # Bioparc Valencia variations
    coming_from %in% c("Bioparc", "BioParc", "Bioparc Valencia", 
                       "Bioparco de Valencia", "Bioparco valencia", 
                       "Bioparco Valencia") ~ "Zoo Bioparc Valencia",

    # Oceanografic Valencia
    coming_from %in% c("Oceanografic Valencia", "Oceanografico") ~ "Aquarium Oceanografic Valencia",

    # Rio Safari Elche variations
    coming_from %in% c("Rio safari Elche", "Rio Safari Elche", 
                       "RIO SAFARI ELCHE", "Zoo safari Elche", 
                       "Zoo Safari Elche","Elche","Elche Safari") ~ "Zoo Rio Safari Elche",

    # Salvi variations
    coming_from %in% c("Salvi", "Salvis") ~ "Salvi",

    # Sendaviva variations
    coming_from %in% c("Sendavia", "Sendaviva") ~ "Zoo Sendaviva Navarra",

    # Torre Ferrusa variations
    coming_from %in% c("Torre Ferrusa (UAB)", "Torrefarrussa", "Torreferrusa") ~ "Zoo Centro de Fauna de Torreferrussa",

    # Zoo Madrid variations
    coming_from %in% c("Zoo de madrid", "Zoo de Madrid", "ZOO de Madrid", 
                       "zoo madrid", "Zoo madrid", "Zoo Madrid", 
                       "ZOO Madrid", "ZOO MADRID","Madrid Zoo") ~ "Zoo Madrid",

    # Zoo Jerez variations
    coming_from %in% c("Zoo Botánico Jerez", "Zoo Jerez", 
                       "Zoologico Jerez de la Frontera","Jerez") ~ "Zoo Jerez",
    
    # Copenhagen zoo
    coming_from %in% c("Copenaghen Zoo","Copenhaguen ZOO") ~ "Zoo Copenhagen",
    
    # Prague zoo
    coming_from %in% c("Praga Zoo") ~ "Zoo Prague",
    
    # "Marcelle Naturaleza"
    coming_from %in% c("Marcelle Naturaleza") ~ "Zoo Marcelle Natureza",
    
    # "Tarragona/CBP"
    
    coming_from %in% c("CBP","Tarragona","Tarragona/CBP")~ "Tarragona/CBP",
    
    # "Aquarium de Madrid"
    coming_from =="Aquarium Madrid"~"Aquarium de Madrid",
    
    # "MondoMar"
    coming_from =="MondoMar"~"Zoo Mundomar Benidorm",
    
    # "Aquarium San Sebastian"
    coming_from =="Aquarium San Sebastian"~"Aquarium de San Sebastian",
    
    # parque de la naturaleza cabarceno
    
    coming_from =="Parque de la Naturaleza Caberceno"~"Zoo Parque de la Naturaleza Cabarceno",
    
    # MCNB
    
    coming_from =="MCNB"~"Museu de Ciencies Naturlas Barcelona",
    
    
    # Exact matches stay as is
    TRUE ~ coming_from
  ))
```


```{r}
work_table_filt<-work_table %>% 
  select("samples","date","animal_species","coming_from_clean","species_360_local_id") %>% 
  distinct() 

work_table_filt<-work_table_filt %>% 
  group_by(samples) %>% 
  mutate(dups=n()) %>% 
  group_by(samples,) %>%
  summarise(date = paste(unique(date), collapse = ", "),
            animal_species =paste(unique(animal_species), collapse = ", "),
            species_360_local_id =paste(unique(species_360_local_id), collapse = ", "),
            coming_from_clean =paste(unique(coming_from_clean), collapse = ", "),
            .groups = "drop") %>% 
  ungroup() %>% 
  group_by(samples) %>% 
  mutate(dups=n()) 

sort(unique(work_table_filt$coming_from_clean))
```
# make a color scale
```{r}
library(scales)
# Assign categories
df <- tibble(name = sort(unique(work_table_filt$coming_from_clean))) %>%
  mutate(category = case_when(
    grepl("(?i)aquarium", name) | grepl("(?i)Oceanografic", name) ~ "Aquarium",
    grepl("(?i)zoo", name) ~ "Zoo",
    TRUE ~ "Other"
  ))

# Generate red->orange->yellow palette
zoo_colors <- colorRampPalette(c("red", "orange", "yellow"))(
  length(unique(df$name[df$category == "Zoo"]))
)


# Aquarium colors: blues
aq_colors  <- scales::hue_pal(h = c(200, 260))(
  length(unique(df$name[df$category == "Aquarium"]))
)

# Other colors: greys
oth_colors <- gray.colors(
  length(unique(df$name[df$category == "Other"])), 
  start = 0.2, end = 0.7
)

color_map <- c(
  setNames(zoo_colors, df$name[df$category == "Zoo"]),
  setNames(aq_colors,  df$name[df$category == "Aquarium"]),
  setNames(oth_colors, df$name[df$category == "Other"])
)
```


```{r}

work_table_filt %>% 
  left_join(combined_sample,by=c("samples"="animal_ID")) %>% 
  mutate(superclass=if_else(is.na(superclass),"Sample_discarded",superclass),
         superclass=factor(superclass,levels=rev(c("Mammalia","Birds","Reptiles","Amphibian","Fish","Sample_discarded")))) %>% 
  ggplot()+
  geom_bar(aes(x=superclass,fill=coming_from_clean))+
  scale_fill_manual(values= color_map)+
  coord_flip()+
  labs(fill="sample source",x="",y="number of samples")

work_table_filt %>% 
  left_join(combined_sample,by=c("samples"="animal_ID")) %>% 
  mutate(superclass=if_else(is.na(superclass),"Sample_discarded",superclass),
         superclass=factor(superclass,levels=rev(c("Mammalia","Birds","Reptiles","Amphibian","Fish","Sample_discarded")))) %>% 
  group_by(coming_from_clean) %>% 
  filter(!is.na(superclass)) %>% 
  mutate(samples=n()) %>% 
  ggplot()+
  geom_bar(aes(x=reorder(coming_from_clean,samples),fill=superclass))+
  coord_flip()+
  labs(fill="class",x="",y="number of samples")

```


The final tables are:
```{r}

df_path<-data.frame(Table=c("Species Table",
"Species Table + Genome Info","Species Table + Genome + Cell lines","Sample Table Mammals","Sample Table other", "Sample Table Fish","Sample Table All"),
Path=c("/output/Species_List.rds","/output/Species_List_with_genome_info.rds","/output/Species_List_with_genome_and_cl_info.rds","/output/Sample_List_Mammal_clean.rds",
 "/output/Sample_List_Tetrapod_clean.rds","/output/Sample_List_fish_clean.rds", "/output/Sample_List_all_species_clean.rds"))

knitr::kable(df_path)

```



