---
title: "Cryozoo overview trees"
output: html_notebook
---

This script was mostly written by claude to generate an overview style tree plot for the cryozoo. With one tree per class based on the Open tree of life topologies


# Biobank Species Overview — Circular Phylogenetic Trees + PhyloPic silhouettes

# Dependencies:
#   install.packages(c("tidyverse", "rotl", "ape", "rphylopic",
#                      "patchwork", "readxl", "RColorBrewer"))
#   BiocManager::install(c("ggtree", "ggtreeExtra"))
#
# rphylopic >= 1.4.0 required (CRAN):  install.packages("rphylopic")

```{r message=FALSE, warning=FALSE}

library(tidyverse)
library(rotl)
library(ape)
library(ggtree)
library(ggtreeExtra)
library(rphylopic)
library(patchwork)
library(RColorBrewer)
library(googlesheets4)

```


# 0.  CONFIGURATION — tweak colours, paths, scaling here

```{r}


# Expected column names in your table :
COL_SPECIES   <- "species_TOL"       # binomial name, e.g. "Homo sapiens"
COL_CLASS     <- "common_group"         # e.g. "Mammalia", "Aves", "Actinopterygii"
COL_ORDER     <- "order"         # e.g. "Carnivora", "Passeriformes"
COL_IUCN      <- "iucn_status"   # LC / NT / VU / EN / CR / DD / NE
COL_WGS       <- "wg_seq"           # TRUE/FALSE or 1/0
COL_RNASEQ    <- "rna_seq"
COL_ATAC      <- "atac_seq"
COL_NANO      <- "wg_nanopore"
COL_REFGENOME <- "genome_match"    # "own" / "closely_related" / "none"

# Minimum species per order to attempt a PhyloPic silhouette
PHYLOPIC_MIN_SPECIES <- 10

# Panel layout: scale panel area by sqrt(n_species)
SIZE_SCALE_FN <- sqrt



# Colour palettes
IUCN_COLOURS <- c(
  LC = "#60C659",
  NT = "#CCE226",
  VU = "#F9E814",
  EN = "#FC7F3F",
  CR = "#D81E05",
  EW = "#542344",
  DD = "#D1D1C6",
  NE = "#AAAAAA"
)

REFGENOME_COLOURS <- c(
  species             = "#2166ac",
  genus = "#92c5de",
  family            = "#f4f4f4"
)

ASSAY_COLOURS <- c(
  wg_seq     = "#7b2d8b",
  rna_seq  = "#e08214",
  atac_seq = "#1b7837",
  wg_nanopore = "grey75"
)

path<-here::here()
```


# 1. Read table with all cryozoo species 

columns: species, class, order, IUCN status

```{r}

species_list<-read_rds(paste0(path,"/output/Species_List_with_genome_and_cl_info.rds")) %>% 
  mutate(common_group=if_else(is.na(class)|class %in%c("Lepidosauria"),"Reptilia",class)) %>% 
  mutate(common_group=if_else(common_group %in%c("Actinopteri","Chondrichthyes"),"Fish",common_group)) %>% 
  mutate(species_TOL=case_match(species,
                                "Astur gentilis"~"Accipiter gentilis",
                                "Caribicus warreni"~"Celestus warreni",
                                .default=species))

#colnames(species_list)
#unique(species_list$species_TOL)

seq_per_sp<-read_sheet("https://docs.google.com/spreadsheets/d/1YI-0bB7OFfjaUM4af9Lv3BAd1X0x9CoCTKZN3BQM4KA/edit?usp=sharing",
            "Sequencing_per_species") %>% 
  janitor::clean_names() %>% 
  mutate(across(2:6,~if_else(.x=="sequenced",T,F))) %>% 
  mutate(across(2:6,~if_else(is.na(.x),F,.)))

species_list<-species_list %>% 
  select(-rna_seq,-wgs_sequencing) %>% 
  left_join(seq_per_sp) %>% 
  mutate(across(79:83,~if_else(is.na(.x),F,.)))
```

```{r}
library(rredlist)

# Check / set your IUCN Red List API token (get one at https://api.iucnredlist.org)

iucn_key <- "DkkVjMp69BoEpZXLicGLfmiHDBDuLdDBSoZG"
iucn_status <- species_list %>%
  mutate(
    iucn_status = map_chr(species, function(sname) {
      parts <- str_split(sname, " ")[[1]]
      res <- tryCatch(
        rl_species(genus = parts[1], species = parts[2], key = iucn_key)$assessments,
        error = function(e) NULL
      )
      if (is.null(res) || !is.data.frame(res) || nrow(res) == 0) return(NA_character_)
      res %>%
        slice_max(assessment_id, n = 1,with_ties = F) %>%
        pull(red_list_category_code)
    })) %>% 
  select(species,iucn_status) %>%
  mutate(iucn_status = iucn_status %>%
    na_if("NA") %>%
    recode(
      `LR/lc` = "LC",
      `V`     = "VU",
      `I`     = "DD"
    ) %>%
    factor(levels = names(IUCN_COLOURS))
  )


unique(iucn_status$iucn_status)

species_list<- species_list %>% 
  left_join(iucn_status)
```


# 2.  PHYLOPIC — fetch and cache silhouettes for qualifying orders


```{r eval=FALSE, include=FALSE}
# Orders with >= PHYLOPIC_MIN_SPECIES across the whole biobank
order_summary <- species_list %>%
  count(
    class = !!sym(COL_CLASS),
    order = !!sym(COL_ORDER),
    name  = "n_species"
  ) %>% 
  filter(n_species >= PHYLOPIC_MIN_SPECIES)

message("Orders qualifying for PhyloPic (>= ", PHYLOPIC_MIN_SPECIES, " species):")
print(order_summary)
```


```{r eval=FALSE, include=FALSE}
# Fetch one silhouette per qualifying order; cache as named list of raster imgs.
# pick_phylopic() finds the best UUID match; get_phylopic() downloads the raster.
fetch_phylopic_safe <- function(order_name) {
  tryCatch({
  uid <- rphylopic::get_uuid(name = order_name)
    img <- rphylopic::get_phylopic(uuid = uid)
    message("  OK: ", order_name)
    img
  }, error = function(e) {
    message("  WARN: PhyloPic not found for '", order_name, "' — ", e$message)
    NULL
  })
}

message("\nFetching PhyloPic silhouettes...")
phylopic_cache <- order_summary %>%
  pull(order) %>%
  set_names() %>%
  map(fetch_phylopic_safe) %>%
  compact()

message(
  length(phylopic_cache), " / ", nrow(order_summary),
  " silhouettes fetched successfully."
)

```



# 3.  HELPER — fetch OTL tree for one class

```{r}

fetch_class_tree <- function(species_vec, class_name) {

  message("  Fetching OTL tree for ", class_name, " (", length(species_vec), " spp)")

  matched <- tryCatch(
    tnrs_match_names(species_vec, context_name = NULL),
    error = function(e) { message("    OTL match failed: ", e$message); NULL }
  )
  if (is.null(matched)) return(NULL)

  ott_ids <- matched %>%
    filter(!is.na(ott_id), !flags %in% c("barren", "extinct")) %>%
    pull(ott_id) %>%
    unique()

  if (length(ott_ids) < 3) {
    message("    Too few resolved names (", length(ott_ids), "), skipping")
    return(NULL)
  }

  tree <- tryCatch(
    tol_induced_subtree(ott_ids = ott_ids, label_format = "name"),
    error = function(e) { message("    Subtree fetch failed: ", e$message); NULL }
  )
  if (is.null(tree)) return(NULL)

  tree$tip.label <- tree$tip.label %>%
    str_replace_all("_", " ") %>%
    str_remove("\\s+ott\\d+$")

  tree <- compute.brlen(tree, method = "Grafen")
  tree <- chronos(tree, quiet = TRUE)
  tree
}

```


# 4.  HELPER — compute angular midpoint for a set of tips in a circular tree
#
#   ggtree places n tips at equal angular spacing.
#   We identify where our order's tips fall and take the circular mean angle,
#   then convert to Cartesian (x, y) at the desired placement radius.

```{r eval=FALSE, include=FALSE}


tips_to_xy <- function(tree, tip_subset, radius) {

  n          <- length(tree$tip.label)
  tip_angles <- tibble(
    label = tree$tip.label,
    # ggtree default: 90° at top, counter-clockwise, in degrees
    angle_deg = 90 - 360 * (seq_len(n) - 1) / n
  )

  subset_ang <- tip_angles %>%
    filter(label %in% tip_subset) %>%
    pull(angle_deg)

  if (length(subset_ang) == 0) return(NULL)

  # Circular mean (handles ±180° wrap-around correctly)
  rad    <- subset_ang * pi / 180
  mean_x <- mean(cos(rad))
  mean_y <- mean(sin(rad))
  mid_rad <- atan2(mean_y, mean_x)

  tibble(
    x         = radius * cos(mid_rad),
    y         = radius * sin(mid_rad),
    angle_deg = mid_rad * 180 / pi
  )
}

```

# 5.  HELPER — build one annotated circular ggtree plot with PhyloPics


```{r}


build_class_plot <- function(tree, meta_df, class_name, phylopic_imgs) {
# tree          = trees[[1]]
# meta_df       = species_list
# class_name    = names(trees[1])
# phylopic_imgs = phylopic_cache
  # Radial offset where silhouettes are placed (beyond outermost ring).
  # 5 annotation rings at offset ~0.08 each push the outer edge to ~1.6;
  # place silhouettes at 2.0 with tip labels reaching ~1.7–1.8.
  SILHOUETTE_RADIUS <- 2.0
  SILHOUETTE_YSIZE  <- 0.12   # height relative to plot coordinates
  LABEL_RADIUS_MULT <- 1.18   # order label placed slightly further out

  tip_df <- tibble(species = tree$tip.label) %>%
    left_join(meta_df, by = c("species" = COL_SPECIES))

  # ── Base circular tree ──────────────────────────────────────────────────────
  p <- ggtree(tree, layout = "circular", size = 0.35, colour = "grey40") +
    geom_tiplab(size = 1.8, offset = 0.05, hjust = 0,
                fontface = "italic", colour = "grey20") +
    labs(title = class_name) +
    theme(
      plot.title      = element_text(face = "bold", size = 11,
                                     hjust = 0.5, margin = margin(b = 4)),
      legend.position = "none"
    )

  # ── Ring 1: IUCN status ─────────────────────────────────────────────────────
  iucn_df <- tip_df %>%
    select(species,all_of(c(COL_IUCN))) %>%
    mutate(iucn_status = factor(iucn_status, levels = names(IUCN_COLOURS)))

  p <- p +
    geom_fruit(
      data        = iucn_df,
      geom        = geom_tile,
      mapping     = aes(y = species, fill = iucn_status),
      width       = 0.08, offset = 0.15,
      colour      = "white", show.legend = FALSE
    ) +
    scale_fill_manual(values = IUCN_COLOURS, na.value = "grey90",
                      name = "IUCN", drop = FALSE)

  # ── Rings 2–4: Assay presence ───────────────────────────────────────────────
  assay_long <- tip_df %>%
    select(species,all_of(c( COL_WGS, COL_RNASEQ, COL_ATAC,COL_NANO))) %>% 
    pivot_longer(-species, names_to = "assay", values_to = "present") %>% 
    mutate(present=if_else(!is.na(present),T,F)) %>% 
    mutate(
      assay    = factor(assay, levels = c("wg_seq", "rna_seq", "atac_seq","wg_nanopore")),
      present  = as.logical(present),
      fill_col = if_else(present, ASSAY_COLOURS[as.character(assay)], NA_character_)
    )

  p <- p +
    geom_fruit(
      data        = assay_long,
      geom        = geom_tile,
      mapping     = aes(y = species, x = assay, fill = fill_col),
      width       = 0.08, offset = 0.05,
      colour      = "white", show.legend = FALSE
    ) +
    scale_fill_identity()

  # ── Ring 5: Reference genome ────────────────────────────────────────────────
  refg_df <- tip_df %>%
    select(species, genome_match) %>%
    mutate(genome_match = factor(genome_match, levels = names(REFGENOME_COLOURS)))

  p <- p +
    geom_fruit(
      data        = refg_df,
      geom        = geom_tile,
      mapping     = aes(y = species, fill = genome_match),
      width       = 0.08, offset = 0.05,
      colour      = "white", show.legend = FALSE
    ) +
    scale_fill_manual(values = REFGENOME_COLOURS, na.value = "grey90",
                      name = "Ref Genome", drop = FALSE)

  # ── PhyloPic silhouettes outside the outermost ring ─────────────────────────
  # Find orders in this class panel that have a cached silhouette
  class_orders <- tip_df %>%
    select(species, order = !!sym(COL_ORDER)) %>%
    filter(order %in% names(phylopic_imgs)) %>%
    distinct()

  if (nrow(class_orders) > 0) {

    # Compute placement position for each qualifying order
    placement_df <- class_orders %>%
      group_by(order) %>%
      summarise(tip_list = list(species), .groups = "drop") %>%
      mutate(
        pos = map(tip_list, ~ tips_to_xy(tree, .x, SILHOUETTE_RADIUS))
      ) %>%
      filter(!map_lgl(pos, is.null)) %>%
      unnest(pos) %>%
      mutate(
        x_lab = x * LABEL_RADIUS_MULT,
        y_lab = y * LABEL_RADIUS_MULT,
        img   = map(order, ~ phylopic_imgs[[.x]])
      )

    # Place each silhouette
    for (i in seq_len(nrow(placement_df))) {
      p <- p +
        add_phylopic_base(
          img   = placement_df$img[[i]],
          x     = placement_df$x[i],
          y     = placement_df$y[i],
          ysize = SILHOUETTE_YSIZE,
          alpha = 0.65,
          color = "grey25"
        )
    }


    # Order name labels just beyond each silhouette
    p <- p +
      geom_text(
        data        = placement_df,
        mapping     = aes(x = x_lab, y = y_lab, label = order),
        size        = 2.0,
        colour      = "grey35",
        fontface    = "italic",
        inherit.aes = FALSE
      )
  }

  p
}

```



# 6.  HELPER — shared legend panel


```{r}

build_legend <- function() {

  p_iucn <- ggplot(
    tibble(x = seq_along(IUCN_COLOURS),
           s = factor(names(IUCN_COLOURS), levels = names(IUCN_COLOURS))),
    aes(x = x, y = 1, fill = s)
  ) +
    geom_tile() +
    scale_fill_manual(values = IUCN_COLOURS, name = "IUCN Status", drop = FALSE) +
    guides(fill = guide_legend(nrow = 1, title.position = "top")) +
    theme_void() +
    theme(legend.position = "right",
          legend.title    = element_text(size = 8, face = "bold"),
          legend.text     = element_text(size = 7))

  p_ref <- ggplot(
    tibble(x = seq_along(REFGENOME_COLOURS),
           g = factor(names(REFGENOME_COLOURS), levels = names(REFGENOME_COLOURS))),
    aes(x = x, y = 1, fill = g)
  ) +
    geom_tile() +
    scale_fill_manual(
      values = REFGENOME_COLOURS,
      name   = "Reference Genome",
      labels = c("Own assembly", "Closely related", "None"),
      drop   = FALSE
    ) +
    guides(fill = guide_legend(nrow = 1, title.position = "top")) +
    theme_void() +
    theme(legend.position = "right",
          legend.title    = element_text(size = 8, face = "bold"),
          legend.text     = element_text(size = 7))

  p_assay <- ggplot(
    tibble(x     = 1:3,
           assay = factor(c("wgs", "rnaseq"),
                          levels = c("wgs", "rnaseq"))),
    aes(x = x, y = 1, fill = assay)
  ) +
    geom_tile() +
    scale_fill_manual(
      values = ASSAY_COLOURS,
      name   = "Sequencing Assay",
      labels = c("WGS", "RNA-seq"),
      drop   = FALSE
    ) +
    guides(fill = guide_legend(nrow = 1, title.position = "top")) +
    theme_void() +
    theme(legend.position = "right",
          legend.title    = element_text(size = 8, face = "bold"),
          legend.text     = element_text(size = 7))

  p_note <- ggplot() +
    annotate("text",
             x = 0.5, y = 0.5,
             label = paste0(
               "Silhouettes shown for orders\n",
               "with \u2265 ", PHYLOPIC_MIN_SPECIES,
               " species (via PhyloPic)"
             ),
             size = 2.8, colour = "grey40", hjust = 0.5) +
    theme_void()

  (p_iucn | p_ref | p_assay | p_note) +
    plot_layout(guides = "keep", widths = c(2, 1.5, 1.5, 1.2))
}


```


# 7.  plot Taxonomies

```{r}

# ── 7a. Class counts ──────────────────────────────────────────────────────────
class_counts <- species_list %>%
  count(class = !!sym(COL_CLASS), name = "n_species") %>%
  
  arrange(desc(n_species))

message("\nSpecies per class:")
print(class_counts)

# ── 7b. Fetch OTL trees ───────────────────────────────────────────────────────
message("\nFetching phylogenetic trees from Open Tree of Life...")

trees <- class_counts %>%
  pull(class) %>%
  set_names() %>%
  map(function(cls) {
    sp_vec <- species_list %>%
      filter(!!sym(COL_CLASS) == cls) %>%
      pull(!!sym(COL_SPECIES))
    fetch_class_tree(sp_vec, cls)
  }) %>%
  compact()

if (length(trees) == 0) stop("No trees could be fetched. Check your species names.")
message("Successfully fetched trees for: ", paste(names(trees), collapse = ", "))

```

```{r}

# ── 7c. Build class plots ─────────────────────────────────────────────────────
message("\nBuilding annotated tree plots...")

class_plots <- trees %>%
  imap(~ build_class_plot(
    tree          = .x,
    meta_df       = species_list,
    class_name    = .y,
    phylopic_imgs = phylopic_cache
  ))

```

```{r}

# ── 7d. Relative panel sizes (sqrt-scaled, largest first) ─────────────────────
sizes_ordered <- class_counts %>%
  filter(class %in% names(class_plots)) %>%
  mutate(size_weight = SIZE_SCALE_FN(n_species)) %>%
  arrange(desc(size_weight)) %>%
  select(class, size_weight) %>%
  deframe()

class_plots   <- class_plots[names(sizes_ordered)]

```


```{r}


# ── 7e. Multipanel layout: 2 rows, larger classes on top ──────────────────────
n_plots      <- length(class_plots)
n_top        <- ceiling(n_plots / 2)

top_plots    <- class_plots[seq_len(n_top)]
bottom_plots <- class_plots[seq(n_top + 1, n_plots)]
top_sizes    <- sizes_ordered[seq_len(n_top)]
bottom_sizes <- sizes_ordered[seq(n_top + 1, n_plots)]

top_row    <- wrap_plots(top_plots,    widths = top_sizes,    nrow = 1)
bottom_row <- wrap_plots(bottom_plots, widths = bottom_sizes, nrow = 1)
legend_row <- build_legend()

combined <- (top_row / bottom_row / legend_row) +
  plot_layout(heights = c(
    max(top_sizes),
    max(bottom_sizes),
    max(top_sizes) * 0.18
  )) +
  plot_annotation(
    title    = "Biobank Species Overview",
    subtitle = paste0(
      "Annotation rings (inside \u2192 outside): ",
      "IUCN status \u2502 WGS \u2502 RNA-seq \u2502 ATAC-seq \u2502 Reference genome\n",
      "PhyloPic silhouettes shown outside outermost ring ",
      "for orders with \u2265 ", PHYLOPIC_MIN_SPECIES, " species"
    ),
    theme = theme(
      plot.title    = element_text(size = 16, face = "bold",    hjust = 0.5),
      plot.subtitle = element_text(size = 8.5, colour = "grey40", hjust = 0.5)
    )
  )

combined
```


```{r}

# ── 7f. Save ──────────────────────────────────────────────────────────────────
ggsave("biobank_overview.pdf", combined,
       width = 24, height = 20, units = "in", device = cairo_pdf)

ggsave("biobank_overview.png", combined,
       width = 24, height = 20, units = "in", dpi = 300)

message("\nDone! Saved to biobank_overview.pdf / .png")

```
