---
title: "3.2 Illumina Project - Species Selection"
output: html_notebook
---

This script aims to select the species for the Illumina project. Run Step 1 (Generate_Lists.Rmd) and Step 2 (Cryo_zoo_summary.Rmd) as well as the selection of species for the silencer project (Tree_selection_silencerproject.Rmd) before.

Lukas: "I favor sampling all available lineages on some level (e.g. family), even if there are no reference available right now, and all available primates"


```{r,message=F,warning=F}


library(tidyverse)
library(cowplot)
library(readxl)
library(googlesheets4)

read_gsheet_allsheets <- function(filename, skip=0, tibble = FALSE) {
  # I prefer straight data.frames
  # but if you like tidyverse tibbles (the default with read_excel)
  # then just pass tibble = TRUE
  sheets <- googlesheets4::sheet_names(filename)[-skip]
  x <- lapply(sheets, function(X) googlesheets4::read_sheet(filename, sheet = X))
  if(!tibble) x <- lapply(x, as.data.frame)
  names(x) <- make.names(sheets)
  x
}

theme_set(theme_classic())

here::here()

path<-here::here()

```

```{r}
List_of_species<-readRDS(paste0(path,"/output/Species_List_with_genome_and_cl_info2.rds")) %>% 
  filter(!duplicated(species))
silencer_species<-readRDS(paste0(path,"/output/selection_silencer_project.rds"))

```


```{r}
# remove fish, select all primates and all that have a genome
selection<-List_of_species %>% 
  filter(superclass != "Fish") %>% 
  filter(fibro_l) %>% 
  mutate(isprimate=order=="Primates") %>% 
  mutate(selection1=cat=="Both"|isprimate)


# select one species for families that are not represented yet

selection2<-selection %>% 
  filter(!selection1) %>%  
  group_by(genus) %>% 
  mutate(s_per_g=n()) %>% 
  group_by(family) %>% 
  slice_max(fibro,n=1) %>%  # select species with most fibroblast lines
  slice_max(individuals,n=1) %>%  # select the one that have the highest number of individuals
  slice_max(s_per_g,n=1) %>%  # select the species with most species per genus
  filter(!(family%in%selection[selection$selection1,]$family)) %>% # remove the ones that are present already in the previous selection
  mutate(selection2=T) %>% 
  select(species,selection2) %>% 
  full_join(selection) %>% 
  mutate(selection1=if_else(is.na(selection1),F,selection1)) %>% 
  mutate(selection2=if_else(is.na(selection2),F,selection2))%>% 
  mutate(illumina= selection1 | selection2) %>% 
  mutate(silencer= species%in%silencer_species$species)


table(selection2$illumina,selection2$genome)
table(selection2$illumina,selection2$superclass)

col_select<-c(colnames(List_of_species)[c(1,2,6,7,9,19,20,21,22,23,24,25,64,65,59,60,67,70,73,29,40)])

final_df<-selection2 %>% 
  ungroup() %>%  
  select(species,illumina,silencer) %>% 
  right_join(List_of_species) %>% 
  select(c(scientific_name,col_select,illumina,silencer)) %>% 
  mutate(illumina=if_else(is.na(illumina),F,illumina)) %>% 
  mutate(silencer=if_else(is.na(silencer),F,silencer))
  

#write_excel_csv(final_df,paste0(path,"/output/all_species_selection.csv"))
#saveRDS(final_df,paste0(path,"/output/all_species_selection.rds"))

selected<-final_df %>% 
  filter(illumina) %>% 
  select(colnames(final_df)[c(1:3,5,9:11,13,20,21)])

#write_excel_csv(selected,paste0(path,"/output/selected_species.csv"))


```

## List of analyses
Lukas: Could you please send a final/frozen list of samples you intend to include and the data types that will be generated for each, including those for which you only want to do the genomes, to make sure we’re on the same page?

```{r}
cryo<-readRDS(paste0(path,"/data/Cryo_stocks_summary.rds")) %>% 
  group_by(animal_ID) %>% 
  filter(type2=="Fibroblast") %>% 
  filter(CT_total>0) %>% 
  select(animal_ID,n_passages,max_passage,CT_total)

sample_table<-readRDS(paste0(path,"/output/Sample_List_all_species_clean.rds")) %>% 
  filter(species %in% final_df$species) %>% 
  filter(cell_type=="Fibro") %>% 
  mutate(cryotube=animal_ID%in%cryo$animal_ID,
         silencer=species%in%subset(final_df,silencer)$species,
         illumina=species%in%subset(final_df,illumina)$species)

df<-sample_table %>% 
  filter(cryotube) %>% 
  group_by(species) %>% 
  summarize(indiv=n()) %>% 
  ungroup() %>% 
  mutate(samp=if_else(indiv>2,2, indiv))

sum(df$samp)


RNA <- read_sheet("https://docs.google.com/spreadsheets/d/1SRid9RGRfcCpGAWsAXG8h0wE-7p1fhOtwvkMJgR1s8w/edit?usp=sharing", 
    sheet = "RNA Extractions",.name_repair = "universal") %>% 
  mutate(Ribonucleic.Acid.ng.ul.=as.numeric(str_extract(Ribonucleic.Acid.ng..U.00B5.l.,"\\d+")))%>% 
  rename(animal_ID=Cryozoo.Number,
         Volume_RNA=ul,
         c_RNA=Ribonucleic.Acid.ng.ul.) %>% 
  select(c(animal_ID,Volume_RNA,c_RNA))


pellet <- read_sheet("https://docs.google.com/spreadsheets/d/1SRid9RGRfcCpGAWsAXG8h0wE-7p1fhOtwvkMJgR1s8w/edit?usp=drive_link", 
    sheet = "Pelet recollit per DNA i RNA",.name_repair = "universal",col_types = c("ccccc")) %>% 
  mutate(pellet=case_when(str_detect(Number.of.vials..80,
                                     pattern="\\s*(RNA|DNA)")~str_count(Number.of.vials..80,pattern="\\s*(RNA|DNA)"),   
                          is.na(Number.of.vials..80)~0,
                          .default = as.numeric(Number.of.vials..80)),
         pellet=if_else(is.na(pellet),1,pellet),
         pellet_type=case_when(str_detect(Number.of.vials..80,pattern="\\s*(RNA&DNA)")~"both",
                               str_detect(Number.of.vials..80,pattern="\\s*(DNA)")~"DNA",
                               str_detect(Number.of.vials..80,pattern="\\s*(RNA)")~"RNA",
                               pellet==0~NA_character_,
                               T~"DNA")) %>% 
  rename(animal_ID=SAMPLE,
         DNA_ex=DNA.extraction.DONE.,
         RNA_ex=RNA.extraction.DONE.) %>%
  left_join(RNA) %>%  
  mutate(DNA_ex=!is.na(DNA_ex),
         RNA_ex=!is.na(RNA_ex)|!is.na(c_RNA)) %>% 
  separate(animal_ID,remove = F,sep = "[^[:alnum:]]+",into = c("animal_ID1","animal_ID2")) %>%
  mutate(animal_ID=if_else(as.numeric(str_remove(animal_ID1,"\\D"))>as.numeric(str_remove(animal_ID2,"\\D"))&!is.na(animal_ID2),animal_ID2,animal_ID1)) %>% 
  mutate(animal_ID2=if_else(animal_ID==animal_ID1,animal_ID2,animal_ID1)) %>% 
  group_by(animal_ID,animal_ID2,pellet_type) %>%
  summarise(DNA_ex=sum(DNA_ex)>0,RNA_ex=sum(RNA_ex)>0, pellet=sum(pellet))

```


```{r}

selected_samples<-left_join(sample_table,pellet) %>% 
  left_join(cryo) %>% 
  mutate(from_cryo=CT_total>4&n_passages>2,
         RNA_ex=if_else(is.na(RNA_ex),F,RNA_ex),
         DNA_ex=if_else(is.na(DNA_ex),F,DNA_ex),
         pellet=if_else(is.na(pellet),0,pellet),
         pellet_RNA=pellet>0&pellet_type%in% c("RNA","both")
         ) %>% 
  group_by(species) %>%
  mutate(selected=case_when(RNA_ex~T,
                            pellet_RNA~T,
                            from_cryo~T,
                            n()<3~T,
                            T~F),
         n_samples=n()) %>% 
  mutate(reps=if_else(n_samples>1,"biorep","techrep1")) %>% 
  mutate(next_step=case_when(RNA_ex~"seq",
                              pellet_RNA~"isolation",
                              from_cryo~"from_cryo",
                              T~"thaw_line"
          ))



 
#write_excel_csv(selected_samples,paste0(path,"/data/selected_celllines_RNA.csv"))
 
selected_samples_short<-selected_samples %>% 
 select(c(animal_ID,sample_id,species,superclass,DNA_ex,RNA_ex,pellet,CT_total,n_passages,reps,next_step))

# write_excel_csv(selected_samples_short,paste0(path,"/data/selected_celllines_RNA_short.csv"))


table(selected_samples$sex)
table(selected_samples$origin)

hist(table(selected_samples$species))

selected_samples_full<-bind_rows(selected_samples,
          selected_samples[which(selected_samples$reps=="techrep1"),] %>% 
            mutate(reps="techrep2"))

hist(table(selected_samples_full$species))

```

```{r}

plot_df<- selected_samples %>% 
  mutate(ispellet=pellet>0) %>% 
  select(species,animal_ID,RNA_ex,from_cryo,silencer,illumina,pellet_RNA) %>% 
  rownames_to_column(var = "index") %>% 
  pivot_longer(cols = c(pellet_RNA,RNA_ex,from_cryo),names_to = "type",values_to = "value") %>% 
  mutate(index=as.numeric(index),
          quarter=case_when(index<nrow(selected_samples)/4~"top",
                           index<nrow(selected_samples)/2~"middle1",
                           index>nrow(selected_samples)/2 & index<(nrow(selected_samples)/4)*3 ~"middle2",
                           T~"bottom"),
         type=factor(type,levels=rev(c("from_cryo","pellet_RNA","RNA_ex")))) 

table(plot_df$quarter)
  

plot_df[is.na(plot_df)]<-F

ggplot(plot_df)+
  geom_tile(aes(x=type,y=reorder(animal_ID,value),fill=value),col="white")+
  facet_wrap(~silencer,ncol=4,scales="free")+
  theme(strip.text = element_blank(),
        strip.background = element_blank(),
        axis.text.x = element_text(angle=90,vjust=0.3))+
  ggsci::scale_fill_npg()+
  xlab("")


table(plot_df$type)
```


```{r}


cryo_long<-readRDS(paste0(path,"/output/Cryo_stocks_long.rds")) %>% 
  filter(type2=="Fibroblast")

cryo_summary<-readRDS(paste0(path,"/output/Cryo_stocks_summary.rds")) %>% 
  filter(type2=="Fibroblast") %>% 
  pivot_longer(cols = 3:(ncol(.)-3),names_to = "passage",values_to = "CT_per_passage") %>% 
  filter(CT_per_passage!=0) %>% 
  mutate(passage=as.numeric(passage)) %>% 
  filter(passage!=0)
  
IDs<-selected_samples_short[selected_samples_short$next_step=="from_cryo",]$animal_ID

positions_pellet<-cryo_long %>% 
  mutate(column=as.numeric(str_extract(column,"\\d+"))) %>% 
  mutate(row=as.numeric(str_extract(row,"\\d+"))) %>% 
  mutate(animal_ID=str_extract(cryotube,pattern = "[Aa]\\d+")) %>% 
  filter(animal_ID%in%IDs) %>% 
  left_join(cryo_summary) %>% 
  group_by(animal_ID) %>% 
  slice_max(CT_per_passage,n = 1) %>% 
  slice_min(passage)

ggplot(positions_pellet)+
  geom_raster(aes(column,row))+
  geom_hline(yintercept = 0.5:10.5)+
  geom_vline(xintercept = 0.5:10.5)+
  scale_y_reverse(breaks=1:10)+
  scale_x_continuous(breaks=1:12,position = "top")+
  geom_text(aes(column,row,label=cryotube),col="white",size=2)+
  facet_wrap(~box)



write_excel_csv(positions_pellet,paste0(path,"/output/positions_pellet.csv"))
```
