---
title: "Illumina Project - Selection"
output: html_notebook
---

Lukas: "I favor sampling all available lineages on some level (e.g. family), even if there are no reference available right now, and all available primates"


```{r,message=F,warning=F}


library(tidyverse)
library(cowplot)
library(readxl)

theme_set(theme_classic())

here::here()

path<-here::here()

```

```{r}
List_of_species<-readRDS(paste0(path,"/output/Species_List_with_genome_and_cl_info2.rds")) %>% 
  filter(!duplicated(species))
silencer_species<-readRDS(paste0(path,"/output/selection_silencer_project.rds"))

```


```{r}
selection<-List_of_species %>% 
  filter(superclass != "Fish") %>% 
  filter(fibro_l) %>% 
  mutate(isprimate=order=="Primates") %>% 
  mutate(selection1=cat=="Both"|isprimate)

selection2<-selection %>% 
  filter(!selection1) %>%  
  group_by(genus) %>% 
  mutate(s_per_g=n()) %>% 
  group_by(family) %>% 
  slice_max(fibro,n=1) %>% 
  slice_max(individuals,n=1) %>% 
  slice_max(s_per_g,n=1) %>% 
  filter(!(family%in%selection[selection$selection1,]$family)) %>% 
  mutate(selection2=T) %>% 
  select(species,selection2) %>% 
  full_join(selection) %>% 
  mutate(selection1=if_else(is.na(selection1),F,selection1)) %>% 
  mutate(selection2=if_else(is.na(selection2),F,selection2))%>% 
  mutate(illumina= selection1 | selection2) %>% 
  mutate(silencer= species%in%silencer_species$species)


table(selection2$illumina,selection2$silencer)
table(selection2$illumina,selection2$superclass)

col_select<-c(colnames(List_of_species)[c(1,2,6,7,9,19,20,21,22,23,24,25,66,65,67,70,75,29,40)])

final_df<-selection2 %>% 
  ungroup() %>%  
  select(species,illumina,silencer) %>% 
  right_join(List_of_species) %>% 
  select(c(scientific_name,col_select,illumina,silencer)) %>% 
  mutate(illumina=if_else(is.na(illumina),F,illumina)) %>% 
  mutate(silencer=if_else(is.na(silencer),F,silencer))
  

write_excel_csv(final_df,paste0(path,"/data/species_selection.csv"))
saveRDS(final_df,paste0(path,"/data/species_selection.rds"))

selected<-final_df %>% 
  filter(illumina) %>% 
  select(colnames(final_df)[c(1:3,5,9:11,13,20,21)])

  write_excel_csv(selected,paste0(path,"/data/selected_species.csv"))

```
## List of analyses
Lukas: Could you please send a final/frozen list of samples you intend to include and the data types that will be generated for each, including those for which you only want to do the genomes, to make sure we’re on the same page?

```{r}

cryo<-readRDS(paste0(path,"/data/Cryo_stocks_long.rds")) %>% 
  group_by(animal_ID) %>% 
  summarize(cryotubes=sum(!is.na(passage))) %>% 
  filter(cryotubes>0)

sample_table<-readRDS(paste0(path,"/output/Sample_List_all_species_clean.rds")) %>% 
  filter(species %in% selected$species) %>% 
  filter(cell_type=="Fibro") %>% 
  filter(animal_ID%in%cryo$animal_ID) %>% 
  left_join(cryo)

df<-sample_table %>% 
  group_by(species) %>% 
  summarize(indiv=n()) %>% 
  ungroup() %>% 
  mutate(samp=if_else(indiv>2,2, indiv))


sum(df$samp)

RNA <- read_excel(paste0(path,"/data/SAMPLES LIST WORK .xlsx"), 
    sheet = "RNA Extractions",.name_repair = "universal") %>% 
  mutate(Ribonucleic.Acid.ng.uL.=as.numeric(str_extract(Ribonucleic.Acid.ng.uL.,"\\d+")))%>% 
  rename(animal_ID=Cryozoo.Number,
         Volume_RNA=ul,
         c_RNA=Ribonucleic.Acid.ng.uL.) %>% 
  select(-c(Sample.Name,Bioanalyzer.result))


pellet <- read_excel(paste0(path,"/data/SAMPLES LIST WORK .xlsx"), 
    sheet = "Pellet recogido para DNA and RN",.name_repair = "universal") %>%
  mutate(pellet=case_when(
    str_detect(Number.of.vials..80,
               pattern="\\s*(RNA|DNA)")~str_count(Number.of.vials..80,pattern="\\s*(RNA|DNA)"),
    is.na(Number.of.vials..80)~0,
    .default = as.numeric(Number.of.vials..80))) %>%
  mutate(pellet=if_else(is.na(pellet),1,pellet)) %>%
  rename(animal_ID=SAMPLE,
         DNA_ex=DNA.extraction.DONE.,
         RNA_ex=RNA.extraction.DONE.) %>% 
  mutate(RNA_ex=if_else(str_detect(RNA_ex,"si")&!is.na(RNA_ex),T,F)) %>% 
  mutate(DNA_ex=if_else(str_detect(DNA_ex,"si")&!is.na(DNA_ex),T,F)) %>% 
  group_by(animal_ID,pellet) %>% 
  summarize(DNA_ex=sum(DNA_ex)>0,
            RNA_ex=sum(RNA_ex)>0
            )



selected_samples<-left_join(sample_table,pellet) %>% 
  left_join(RNA)

table(duplicated(selected_samples$species),selected_samples$superclass)



```


# Plot availability of samples
```{r}

length(unique(selected_samples$species))


select_mat<-selected_samples %>% 
  select(species, animal_ID, cryotubes, pellet, RNA_ex, DNA_ex) %>% 
  mutate(cryotubes=cryotubes>2) %>% 
  mutate(pellet=pellet>0) %>% 
  filter(!duplicated(animal_ID)) %>% 
  pivot_longer(c(cryotubes, pellet, RNA_ex, DNA_ex),names_to = "type",values_to = "value") %>%
  mutate(type=factor(type,levels=c("cryotubes","pellet","DNA_ex","RNA_ex")))

select_mat$half<-paste(rep(c("upper","middle","lower","bottom"),each=nrow(select_mat)/4))

select_mat[is.na(select_mat)]<-F

p<-ggplot(select_mat, aes(y = animal_ID, x = type, fill = value)) +
  geom_tile(color = "white",
            lwd = 0.5,
            linetype = 1) +
  facet_wrap(~half,scales = "free",ncol = 4)+
  theme(
  strip.background = element_blank(),
  strip.text.x = element_blank(),
  axis.text.x = element_text(angle = 90, vjust= 0.5))
p


```

