---
title: "Make Sample and Species Tables"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

In this script the tables from the Google sheet *List of Species* will be summarized into one species table and one sample table. The species table contains all the information on a species level such as, genome avaliability, iucn status, taxonomy and number and type of samples. The sample table will contain information about the individual samples such as sex, age, tissue of origin, storage location etc.

Those tables will be generated for mammals as well as for the remaining tetrapods (Birds,Reptiles, Amphibians).


```{r,message=F,warning=F}

library(tidyverse)
library(cowplot)
library(readxl)
library(runner)

theme_set(theme_classic())

here::here()

path<-here::here()

```
# Getting Species level table
## Load all species summary table
Import table:
```{r}
List_of_species <- read_excel(paste0(path,"/data/List of species.xlsx"), 
    sheet = "SPECIES LIST")

colnames(List_of_species)<-janitor::make_clean_names(colnames(List_of_species))

List_of_species<-List_of_species %>% 
    rename(specie_name_cites=scientific_name)
```

Tidy up table:
```{r}
List_of_species<-List_of_species %>% 
  filter(class!="FISH") %>% 
  mutate(class=str_to_title(class),
         class=if_else(class=="Bird","Aves",class)) %>% 
  mutate(cell_line=gsub(cell_line,pattern = "x",replacement = "Fibro")) %>% 
  mutate(cell_line=gsub(cell_line,pattern = "X",replacement = "Fibro")) %>% 
  mutate(cell_line=gsub(cell_line,pattern = "PBMc/Fibro",replacement = "Fibro/PBMc"))%>%
  mutate(cell_line=if_else(is.na(cell_line),"No cell line",cell_line)) %>% 
  mutate(cell_line=if_else(is.na(cell_line),"No cell line",cell_line)) %>% 
  mutate(iucn_category=if_else(iucn_category %in% c(NA,"NOT APPLICABLE","DD","??"),"NA",iucn_category))%>% 
   mutate(iucn_category=factor(iucn_category,levels=c("NA","LEAST CONCERN","VULNERABLE","NEAR THREATENED","ENDANGERED","CRITICALLY ENDANGERED","EXTINCT IN THE WILD"))) %>% 
  mutate(samples=as.numeric(samples),
         samples=if_else(is.na(samples),1,samples)) %>% 
  mutate(order=as.factor(order)) %>% 
  separate(specie_name_cites,into=c("n1","n2","subspecies")) %>% 
  mutate(n1=stringr::str_to_title(n1),
         n2=tolower(n2)) %>% 
  unite(c(n1,n2),col = "specie_name_cites",sep=" ") %>% 
  mutate(specie_name_cites=gsub(x=specie_name_cites,pattern = "?",replacement = "")) %>%  
  arrange(order) 

```
Replace problematic species names
```{r}
species_names<-read_excel(paste0(path,"/data/Wrong_Species_names.xlsx"))

species_names<-species_names[match(List_of_species$specie_name_cites[
  which(List_of_species$specie_name_cites%in%species_names$specie_name_cites_species)],
  species_names$specie_name_cites_species,
  incomparables = F),] %>% 
  mutate(specie_name_cites=specie_name_cites_species)

List_of_species<-List_of_species %>% 
  left_join(species_names) %>%
  mutate(specie_name_cites=if_else(!(is.na(goat_name)),
                                   goat_name,
                                   specie_name_cites),
         alternative_name=if_else(specie_name_cites==goat_name,
                                   specie_name_cites_species,
                                   specie_name_cites)) %>%  
  select(-c(goat_name,specie_name_cites_sample,specie_name_cites_species,group))



List_of_species[duplicated(List_of_species$specie_name_cites),]
```

There are duplicated species in the species table which turn out to be subspecies. I will treat them as the same for now.

```{r}
dup_sum<-full_join(List_of_species[duplicated(List_of_species$specie_name_cites),],
           List_of_species[duplicated(List_of_species$specie_name_cites,fromLast = T),],
           by=c("specie_name_cites",
               "family",
               "subfamily",
               "cites_annex",
               "class")) %>% 
  mutate(subspecies=case_when(is.na(subspecies.x)~subspecies.y,
                              is.na(subspecies.y)~subspecies.x,
                              !(is.na(subspecies.x))&!(is.na(subspecies.y))~paste(subspecies.x,
                                                                                  subspecies.y,sep=",")),
         common_name=case_when(is.na(common_name.x)~common_name.y,
                              is.na(common_name.y)~common_name.x,
                              !(is.na(common_name.x))&!(is.na(common_name.y))~paste(common_name.x,
                                                                                  common_name.y,sep=",")),
         cell_line=case_when(is.na(cell_line.x)~cell_line.y,
                              is.na(cell_line.y)~cell_line.x,
                              !(is.na(cell_line.x))&!(is.na(cell_line.y))~paste(cell_line.x,
                                                                                  cell_line.y,sep=",")),
         sample=case_when(is.na(sample.x)~sample.y,
                              is.na(sample.y)~sample.x,
                              !(is.na(sample.x))&!(is.na(sample.y))~paste(sample.x,
                                                                                  sample.y,sep=",")),
         samples=samples.x+samples.y,
         order=order.x,
         iucn_category=iucn_category.x,
         karyotype=NA,
         dna=NA,
         rna=NA,
         sequencing=NA,
         i_psc=NA
         ) %>% 
  select(-c(ends_with(".x"),ends_with(".y")))
  
dup_species<-List_of_species[duplicated(List_of_species$specie_name_cites),]$specie_name_cites

List_of_species<-List_of_species %>% 
  filter(!(specie_name_cites %in% dup_species)) %>% 
  bind_rows(dup_sum) %>% 
  mutate(order=as.character(order)) %>% 
  mutate(scientific_name_=gsub(specie_name_cites,pattern = " ",replacement = "_"))

```

```{r}
head(List_of_species)

saveRDS(List_of_species,paste0(path,"/output/Species_List.rds"))
```


## Add genome information form Genome on a tree

```{r}
GOAT_info <- read_delim(paste0(path,"/data/GenomeOnATree_tetra_3.10.22.tsv"), 
    delim = "\t", escape_double = FALSE, 
    col_types = cols(taxon_id = col_character()), 
    trim_ws = TRUE)
  

table((List_of_species$specie_name_cites %in% GOAT_info$scientific_name),List_of_species$class)

List_of_species[!(List_of_species$specie_name_cites %in% GOAT_info$scientific_name),]

List_of_species_new<-inner_join(List_of_species,GOAT_info,by=c("specie_name_cites"="scientific_name")) %>% 
  mutate(common_name=case_when(is.na(common_name.x)~common_name.y,
                               common_name.x==common_name.y~common_name.y,
                               common_name.x!=common_name.y~paste(common_name.x,common_name.y,sep=","),
                               T~NA_character_),
         family=case_when(is.na(family.y)~family.x,
                          is.na(family.x)~family.y,
                               family.x==family.y~family.y,
                               family.x!=family.y~family.y,
                               T~NA_character_),
         class=case_when(is.na(class.y)~class.x,
                         is.na(class.x)~class.y,
                         class.x==class.y~class.y,
                         class.x!=class.y~class.x,
                         T~NA_character_),
         order=case_when(is.na(order.y)~order.x,
                         is.na(order.x)~order.y,
                               order.x==order.y~order.y,
                               order.x!=order.y~order.y,
                               T~NA_character_)) %>%
  select(-c(ends_with(".x"),ends_with(".y"),taxon_rank))


saveRDS(List_of_species_new,paste0(path,"/output/Species_List_with_genome_info.rds"))

```

# Getting Sample level tables
## Sample table Mammals

```{r}

mammals_list <- read_excel(paste0(path,"/data/List of species.xlsx"), 
    sheet = "MAMMALIA", skip = 1)[-1,-2]

colnames(mammals_list)<-janitor::make_clean_names(colnames(mammals_list))


##Make a clean tidy table for mammals
 

mammals_list_clean<-mammals_list %>% 
  mutate(x1=fill_run(x1,run_for_first = F),
         specie_name_cites=fill_run(specie_name_cites),
         common_name=fill_run(common_name),
         family=fill_run(family),
         iucn=fill_run(iucn)) %>% 
  filter(!(is.na(zims_species360)&is.na(barcode))) %>% 
  mutate(barcode=if_else(is.na(barcode)& grepl(x = x7,pattern = "^A"),x7,barcode)) %>%  
  mutate(x7=if_else(grepl(x = x7,pattern = "^A"),NA_character_,x7)) %>%  
  mutate(iPSC=grepl(x7,pattern="iPSC")) %>% 
  separate(barcode,sep = "/",into = c("position1","position2","position3")) %>% 
  separate(x7,sep = "/",into = c("cell_type","type1","type2","type3","type4")) %>% 
  mutate(cell_type=case_when(cell_type=="C"~"Fibro",
                             cell_type=="PBMC"~"PBMc",
                             cell_type=="BLOOD"~"PBMc",
                             cell_type=="PMBc isolation"~"PBMc",
                             cell_type=="?"~"unclear",
                             cell_type=="-"~"unclear",
                             is.na(cell_type)~"Primary",
                             T~as.character(cell_type))) %>% 
  mutate(status=cell_type!="Primary") %>% 
  separate(specie_name_cites,into=c("n1","n2","subspecies")) %>% 
  mutate(n1=stringr::str_to_title(n1),
         n2=tolower(n2)) %>% 
  unite(c(n1,n2),col = "specie_name_cites",sep=" ") %>% 
  mutate(specie_name_cites=gsub(x=specie_name_cites,pattern = "?",replacement = ""),
         class="Mammalia") 

species_names_mammals<-subset(species_names,group=="Mammals")

species_names_mammals<-species_names_mammals[match(mammals_list_clean$specie_name_cites[
  which(mammals_list_clean$specie_name_cites%in%species_names_mammals$specie_name_cites_sample)],
  species_names_mammals$specie_name_cites_sample,
  incomparables = F),]

mammals_list_clean$specie_name_cites[which(
  mammals_list_clean$specie_name_cites%in%species_names_mammals$specie_name_cites_sample)
  ]<-species_names_mammals$goat_name

saveRDS(mammals_list_clean,paste0(path,"/output/Sample_List_Mammal_clean.rds"))
write_excel_csv(mammals_list_clean,paste0(path,"/output/Sample_List_Mammal_clean.csv"))


```


## Sample Table other Tetrapods

```{r}

tetra_list <- read_excel(paste0(path,"/data/List of species.xlsx"), 
    sheet = "AMPHIBIAN",
    col_names = T)[-1,-2]

tetra_list$class<-"Amphibian"

tetra_list<-bind_rows(tetra_list,
                      read_excel(paste0(path,"/data/List of species.xlsx"), 
    sheet = "REPTIL",col_names = T)[-c(1),-2])

tetra_list$class[is.na(tetra_list$class)]<-"Reptiles"

tetra_list<-bind_rows(tetra_list,
                      read_excel(paste0(path,"/data/List of species.xlsx"), 
    sheet = "BIRD",col_names = T,skip = 1)[-1,-2])


tetra_list$class[is.na(tetra_list$class)]<-"Aves"

colnames(tetra_list)<-janitor::make_clean_names(colnames(tetra_list))

```

```{r}


tetra_list_clean<-tetra_list %>% 
  mutate(x1=fill_run(x1,run_for_first = F),
         specie_name_cites=fill_run(specie_name_cites),
         common_name=fill_run(common_name),
         family=fill_run(family),
         iucn=fill_run(iucn)) %>% 
  filter(!(is.na(zims_species360)&is.na(x7)&is.na(x6))) %>% 
  rename(barcode=x7) %>% 
  separate(barcode,sep = "/",into = c("position1","position2","position3")) %>% 
  separate(x6,sep = "/",into = c("cell_type","type1","type2","type3","type4")) %>% 
  mutate(cell_type=case_when(cell_type=="C"~"Fibro",
                             cell_type=="PBMC"~"PBMc",
                             cell_type=="BLOOD"~"PBMc",
                             cell_type=="PMBc isolation"~"PBMc",
                             cell_type=="?"~"unclear",
                             cell_type=="-"~"unclear",
                             is.na(cell_type)~"Primary",
                             T~as.character(cell_type))) %>% 
  mutate(status=cell_type!="Primary") %>% 
  separate(specie_name_cites,into=c("n1","n2","subspecies")) %>% 
  mutate(n1=stringr::str_to_title(n1),
         n2=tolower(n2)) %>% 
  unite(c(n1,n2),col = "specie_name_cites",sep=" ") %>% 
  mutate(specie_name_cites=gsub(x=specie_name_cites,pattern = "?",replacement = "")) %>% 
  mutate(sample_id=NA,
         iPSC=F)

### correct names
species_names_tetra<-subset(species_names,group!="Mammals")

species_names_tetra<-species_names_tetra[match(tetra_list_clean$specie_name_cites[
  which(tetra_list_clean$specie_name_cites%in%species_names_tetra$specie_name_cites_sample)],
  species_names_tetra$specie_name_cites_sample,
  incomparables = F),]

tetra_list_clean$specie_name_cites[which(
  tetra_list_clean$specie_name_cites%in%species_names_tetra$specie_name_cites_sample)
  ]<-species_names_tetra$goat_name


saveRDS(tetra_list_clean,paste0(path,"/output/Sample_List_Tetrapod_clean.rds"))
write_excel_csv(tetra_list_clean,paste0(path,"/output/Sample_List_Tetrapod_clean.csv"))

```

## Combine sample tables

```{r}
combined_sample<-bind_rows(tetra_list_clean,mammals_list_clean) %>% 
  select(-x1) %>% 
  mutate(species_short=paste(str_to_title(substr(str_split(string = specie_name_cites,
                                                           pattern = " ",
                                                           simplify=T)[,1],1,3)),
                             str_to_title(substr(str_split(string = specie_name_cites,
                                                           pattern = " ",
                                                           simplify=T)[,2],1,3)),sep="_")) %>% 
  group_by(species_short) %>% 
  mutate(sample_id=paste(sep="_",species_short,row_number()))
  
saveRDS(combined_sample,
       paste0(path,"/output/Sample_List_all_species_clean.rds"))
write_excel_csv(combined_sample,
                paste0(path,"/output/Sample_List_all_species_clean.csv"))
```

Add Info about cell lines to Species Table:

```{r}
summary_species<-combined_sample %>% 
  group_by(specie_name_cites,subspecies,common_name,family) %>% 
  summarize(samples_cl=n(),
         cell_lines=sum(status),
         fibro=sum(cell_type=="Fibro"),
         pbmc=sum(cell_type=="PBMc"),
         ipsc=sum(iPSC)) %>% 
ungroup()

Species_list_complete<-List_of_species_new%>% 
  inner_join(select(summary_species,c(specie_name_cites,samples_cl,cell_lines,fibro,pbmc,ipsc)),by="specie_name_cites")

saveRDS(Species_list_complete,paste0(path,"/output/Species_List_with_genome_and_cl_info.rds"))

```


## Compare sample list and species list info

Entries only found in sample table: `r setdiff(List_of_species$specie_name_cites,summary_species$specie_name_cites)`

Entries only found in sample table:`r setdiff(summary_species$specie_name_cites,List_of_species$specie_name_cites)`


```{r}
List_of_species %>% 
  select(specie_name_cites,samples,sample,sex,cell_line,i_psc) %>% 
  inner_join(summary_species) %>% 
  ggplot()+
  geom_point(aes(samples,samples_cl))+
  geom_abline(slope = 1,intercept = 0)

sum<-List_of_species %>% 
  select(specie_name_cites,samples,sample,sex,cell_line,i_psc) %>% 
  inner_join(summary_species) %>% 
  mutate(comp=case_when(samples==samples_cl~"same",
                        samples>samples_cl~"more in species list",
                        samples<samples_cl~"more in sample list"
                        ))

table(sum$comp)

```


The final tables are:
```{r}

df_path<-data.frame(Table=c("Species Table",
"Species Table + Genome Info","Species Table + Genome + Cell lines","Sample Table Mammals","Sample Table other","Sample Table All"),
Path=c("/output/Species_List.rds","/output/Species_List_with_genome_info.rds","/output/Species_List_with_genome_and_cl_info.rds","/output/Sample_List_Mammal_clean.rds",
 "/output/Sample_List_Tetrapod_clean.rds", "/output/Sample_List_all_species_clean.rds"))

knitr::kable(df_path)

```



