---
title: "2. Plotting the Cryozoo"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r,message=F,warning=F}


library(tidyverse)
library(cowplot)
library(readxl)

theme_set(theme_classic())

here::here()

path<-here::here()

# load colour palettes

source("https://raw.githubusercontent.com/mammalian-iPSCs/colour-palettes/main/palettes.R")

```
This script aims to summarize and plot the species present in the cryozoo. It is based on lists from the "Clean_CryoZoo_Lists.Rmd" script so make sure to run this before.

```{r}
List_of_species<-readRDS(paste0(path,"/output/Species_List_with_genome_and_cl_info.rds"))


table(List_of_species$fibro_l)

sum_list<-List_of_species %>% 
  group_by(superclass) %>% 
  summarize(nspecies=n())

sum(sum_list$nspecies)


table(List_of_species$superclass)

```
## General Species level summary

```{r, fig.dim=c(8,7)}

bar.all_spec<-ggplot(data=List_of_species)+
  geom_bar(aes(x="All\n Samples",fill=superclass))+
  geom_text(data=sum_list,aes(label=nspecies),x="All\n Samples",y=c(320,300,255,175,60),color="white")+
  labs(x="",y="Number of species",fill="")+
  coord_flip()+
  theme_minimal()+
  theme(legend.position = "top",
        axis.ticks.y=element_blank(),
        axis.line.y = element_blank())+
  scale_fill_manual(values=class_col,limits=force)



bar.fibro<-ggplot(data=List_of_species)+
  geom_bar(aes(x=superclass,fill=superclass,alpha=fibro_l))+
  labs(x="",y="Number of species",fill="",alpha="Fibroblasts established")+
  coord_flip()+
  ggsci::scale_fill_aaas()+
  theme_minimal()+
  theme(legend.position = c(0.8,0.4),
        axis.ticks.y=element_blank(),
        axis.line.y = element_blank())+
  guides(fill="none")+
  scale_fill_manual(values=class_col,limits=force)


plot_grid(bar.all_spec,bar.fibro,nrow = 2,rel_heights = c(0.3,0.7),align = "v",axis="tblr")


save_plot(plot = plot_grid(bar.all_spec,bar.fibro,nrow = 2,rel_heights = c(0.3,0.7),align = "v",axis="tblr"),base_height = 5,base_width = 8,filename = paste0(path,"/output/figures/overview.png"))

```

## Cell lines per order

```{r}

taxlevel<-List_of_species %>% 
  group_by(order) %>% 
  mutate(samples_per_taxlevel=n()) %>% 
  group_by(class) %>% 
  mutate(taxlevel=if_else(samples_per_taxlevel>5,"order","class"),
        samples_per_taxlevel=if_else(samples_per_taxlevel>5,samples_per_taxlevel,n()),
        taxon=if_else(taxlevel=="order",order,class),
        taxon=if_else(is.na(taxon),superclass,taxon)) %>% 
  select(order,class,samples_per_taxlevel,taxlevel,taxon) %>% 
  distinct()
  
  

for (i in unique(taxlevel$taxon)) {

print(i)
tl<-taxlevel %>% 
  filter(taxon==i) %>% 
  select(taxlevel) %>% 
  distinct() %>% 
  pull(taxlevel)

filt <- List_of_species %>% 
  filter(if (tl == "order") order == i else class == i)

filt %>% 
ggplot()+
  geom_bar(aes(x=family,fill=superclass,alpha=fibro_l))+
  labs(x="",y="Number of species",fill="",alpha="Fibroblasts?")+
  coord_flip()+
  theme_minimal()+
  theme(legend.position = "right",
        axis.ticks.y=element_blank(),
        axis.line.y = element_blank())+
  guides(fill="none")+
  scale_fill_manual(values=class_col,limits=force)+
  ggtitle(paste(i,"Species with Fibroblast lines"))->p
  
print(p)
}

```

## Species by IUCN status
```{r}


names(IUCN_COLOURS) <- c("LEAST CONCERN", "NEAR THREATENED", "VULNERABLE", "ENDANGERED",
                          "CRITICALLY ENDANGERED", "EXTINCT IN THE WILD", "EXTINCT",
                          "DATA DEFICIENT", "NOT EVALUATED")

bar.iucn<-List_of_species %>% 
  filter(!is.na(iucn_category)) %>% 
  mutate(iucn_category=factor(iucn_category,levels=c("EXTINCT IN THE WILD","CRITICALLY ENDANGERED","ENDANGERED","VULNERABLE","NEAR THREATENED","LEAST CONCERN","NA")
                              )) %>% 
  ggplot()+
  geom_bar(aes(x=superclass,fill=iucn_category),position=position_stack())+
  labs(x="",y="Number of species",fill="IUCN status")+
  coord_flip()+
  theme_minimal()+
  theme(legend.position = c(0.7,0.3),
        axis.ticks.y=element_blank(),
        axis.line.y = element_blank())+
  scale_fill_manual(values=IUCN_COLOURS)
bar.iucn

  
save_plot(plot = bar.iucn,base_height = 5,base_width = 5,filename = paste0(path,"/output/figures/bar.iucn.png"))

```

## IUCN status all

```{r}

List_of_species %>% 
  filter(!is.na(iucn_category)) %>% 
  mutate(iucn_category=factor(iucn_category,levels=c("EXTINCT IN THE WILD","CRITICALLY ENDANGERED","ENDANGERED","VULNERABLE","NEAR THREATENED","LEAST CONCERN","NA")
                              )) %>% 
  ggplot()+
  geom_bar(aes(x="all",fill=iucn_category),position=position_stack())+
  labs(x="",y="Number of species",fill="IUCN status")+
  theme_minimal()+
  scale_fill_manual(values=IUCN_COLOURS)

table(List_of_species$iucn_category)

```

## Replicates per Class

```{r,fig.dim=c(10,12)}

p.samples<-ggplot(data=List_of_species)+
  geom_bar(aes(x=individuals,fill=superclass),position = position_dodge(),show.legend = F)+
  labs(x="Number of individuals",y="Number of species",fill="")+
  facet_grid(superclass~.)+
  scale_fill_manual(values=class_col,limits=force)+
  scale_x_continuous(breaks=c(1:20))+
  theme_bw()

p.samples

save_plot(plot = p.samples,base_height = 5,base_width = 5,filename = paste0(path,"/output/figures/p.samples.png"))

```

## Replicates binned

```{r}

p.samples2<-List_of_species %>% 
  mutate(samples_cat=case_when(individuals==1~ "one",
                               individuals==2~ "two",
                               individuals==3 ~ "three",
                               individuals>3 ~ "more than three")) %>% 
  mutate(samples_cat=factor(samples_cat,levels=rev(c("more than three","three","two","one")))) %>% 
  mutate(superclass=forcats::fct_rev(superclass)) %>% 
  
    ggplot()+
  geom_bar(aes(superclass,fill=samples_cat),position = position_dodge())+
  coord_flip()+
  scale_fill_viridis_d(direction = -1)+
  labs(y="number of species",x="",fill="Replicates per species")+
  theme(legend.position = "top")

p.samples2

save_plot(plot = p.samples2,base_height = 5,base_width = 5,filename = paste0(path,"/output/figures/p.samples2.png"))

```



# representation of cryozoo species

```{r}

all_species_summary<-readRDS(paste0(path,"/data/mammal_data_orders.rds"))

```



```{r}
ggplot(all_species_summary)+
  geom_col(aes(y=nspecies,x=group),position = position_dodge())+
  facet_wrap(~order,scales = "free_x")+
  coord_flip()
```


```{r}
ggplot(all_species_summary)+
  geom_col(aes(y=ngenera,x=group),position = position_dodge())+
  facet_wrap(~order,scales = "free_x")+
  coord_flip()
```


```{r}
ggplot(all_species_summary)+
  geom_col(aes(y=nfamily,x=group),position = position_dodge())+
  facet_wrap(~order)+
  coord_flip()
```


```{r}
all_mammals<-all_species_summary %>% 
  filter(group=="all_mammals") %>% 
  ungroup() %>% 
  select(-c(group)) %>% 
  rename(total_genera=ngenera,
         total_species=nspecies,
         total_families=nfamily)
  
 all_species_summary<-all_species_summary %>% 
   filter(group!="all_mammals") %>% 
   left_join(all_mammals) %>% 
   mutate(p_species=(nspecies/total_species)*100,
          p_genera=(ngenera/total_genera)*100,
          p_families=(nfamily/total_families)*100)
    
  ggplot()+
    ggbeeswarm::geom_beeswarm(data=all_species_summary,aes(x=reorder(order,total_species),col=group,y=p_species,size=total_species))+
    coord_flip()+
    theme_light()+
      scale_colour_manual(values=c("zoonomia"="#e63946","cryozoo"="#6987b5"))+
      labs(x="mammalian orders\n (sorted by number of species)",y="percent of all species covered")
```


```{r}
sum_long<-all_species_summary %>% 
  pivot_longer(cols=c(p_species,p_genera,p_families),names_to = c("P","tax_level"),values_to = "percent",names_sep = "_") %>% 
  select(-P) %>% 
  mutate(tax_level=factor(tax_level,levels=rev(c("families","genera","species"))))

ggplot()+
    ggbeeswarm::geom_beeswarm(data=sum_long,aes(x=reorder(order,total_species),colour=group,y=percent),position=position_dodge())+
    coord_flip()+
    theme_light()+
      scale_colour_manual(values=c("zoonomia"="#e63946","cryozoo"="#6987b5"))+
      labs(x="mammalian orders\n (sorted by number of species)",y="percent of all taxonomic units covered") +
  facet_grid(~tax_level)
  
```


```{r}

taxlevel_colours<- c(
    "species" = "#2AA37F",
    "genera"   = "#E18445",
    "families"  = "#9490C3"
  )

sum_long<-all_species_summary %>% 
  filter(group=="cryozoo") %>% 
  pivot_longer(cols=c(p_species,p_genera,p_families),names_to = c("P","tax_level"),values_to = "percent",names_sep = "_") %>% 
  select(-P) %>% 
  mutate(tax_level=factor(tax_level,levels=rev(c("families","genera","species"))))

ggplot()+
    ggbeeswarm::geom_beeswarm(data=sum_long,aes(x=reorder(order,total_species),colour=tax_level,y=percent),position=position_dodge())+
    coord_flip()+
    theme_light()+
      scale_colour_manual(values=taxlevel_colours)+
      labs(x="mammalian orders\n (sorted by number of species)",y="percent of all taxonomic units covered")
  
```

# Summary plots genome

## Refgenome sequencing 
```{r}
# per superclass

List_of_species%>% 
  mutate(sequencing_status=factor(sequencing_status,levels=rev(c("sample_collected","sample_acquired","in_progress","open","insdc_open","published")))) %>% 
  mutate(superclass=factor(superclass,levels=rev(c("Mammalia","Birds","Amphibia","Reptil","Fish")))) %>% 

ggplot()+
  geom_bar(aes(fill=sequencing_status,x=superclass))+
  coord_flip()
```


```{r}
# per order

List_of_species %>% 
  mutate(sequencing_status=factor(sequencing_status,levels=rev(c("sample_collected","sample_acquired","in_progress","open","insdc_open","published")))) %>% 
   mutate(superclass=factor(superclass,levels=rev(c("Mammalia","Birds","Amphibia","Reptil","Fish")))) %>% 
ggplot()+
  geom_bar(aes(fill=sequencing_status,x=order))+
  coord_flip()+
  facet_grid(superclass~genome,scales = "free",space = "free") 



```


# Only Fibroblast cell lines:

### Based on species table
```{r}
List_of_cell_lines<-List_of_species %>% 
  filter(fibro_l) %>% 
  group_by(order) %>% 
  mutate(nspecies=n()) %>% 
  ungroup() %>% 
  arrange(nspecies)

order_col <- scales::hue_pal()(length(unique(List_of_cell_lines$order)))
names(order_col)<-unique(List_of_cell_lines$order)


p.species_per_order<-ggplot(List_of_cell_lines)+
  geom_bar(aes(y=reorder(order,nspecies),fill=order),show.legend = F)+
  labs(y="",x="Number of species")+
  scale_x_reverse()+
  theme(axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.line.y = element_blank())+
  scale_fill_manual(values=order_col,limits=force)+
  facet_grid(superclass~.,scales = "free",space="free")


p.sample_per_species<-ggplot(List_of_cell_lines)+
  stat_summary(fun.data = "mean_se",aes(x=individuals,y=reorder(order,nspecies),col=order),show.legend = F)+
  #ggbeeswarm::geom_quasirandom(aes(x=individuals,y=reorder(order,nspecies),colour=order),groupOnX = F,show.legend = F)+
  labs(y="",x="Number of samples")+
  scale_x_continuous(breaks=1:15)+
  scale_colour_manual(values=order_col,limits=force)+
  facet_grid(superclass~.,scales = "free",space="free")


p.byorder<-plot_grid(p.species_per_order,p.sample_per_species,rel_widths = c(0.45,0.55))

p.byorder

save_plot(plot = p.byorder,base_height = 8,base_width = 8,filename = paste0(path,"/output/figures/p.byorder.png"))

```



```{r}

library(colorspace)


make_shaded_palette <- function(base_cols, shade_labels, amount = 0.3, name_sep=" ", direction=c("light_to_dark", "dark_to_light")) {
  direction <- match.arg(direction)
  n_shades <- length(shade_labels)
  
  # spread shade amounts evenly from +amount (lightest) to -amount (darkest)
  # e.g. n=2 -> c(0.3, -0.3); n=3 -> c(0.3, 0, -0.3); n=5 -> c(0.3, 0.15, 0, -0.15, -0.3)
  shade_amounts <- seq(amount, -amount, length.out = n_shades)
  
  if (direction == "dark_to_light") {
    shade_amounts <- rev(shade_amounts)
  }
  
  shaded <- lapply(names(base_cols), function(nm) {
    sapply(shade_amounts, function(a) {
      if (a > 0) lighten(base_cols[[nm]], amount = a)
      else if (a < 0) darken(base_cols[[nm]], amount = abs(a))
      else base_cols[[nm]]
    })
  })
  
  
  
  shaded_cols <- unlist(shaded)
  names(shaded_cols) <- as.vector(outer(shade_labels, names(base_cols),
                                         FUN = function(s, n) paste(n, s, sep = name_sep)))
  
  # reorder so it's grouped by base colour, not by shade
  shaded_cols <- shaded_cols[order(rep(seq_along(base_cols), each = n_shades))]
  
  shaded_cols
}

class_genome_cols <- make_shaded_palette(class_col, 
                                         shade_labels = c("Not Available","Available"),
                                         direction="light_to_dark")

List_of_species %>% 
  mutate(genome_class=if_else(genome,paste(superclass, "Available"),paste(superclass,"Not Available")),
         genome_class=factor(genome_class,levels = rev(c("Mammalia Available", "Mammalia Not Available","Birds Available","Birds Not Available","Amphibia Available","Amphibia Not Available","Reptil Available","Reptil Not Available","Fish Available","Fish Not Available")))) %>% 
ggplot()+
  geom_bar(aes(fill=genome_class,x=superclass),position=position_fill())+
  coord_flip()+
  labs(y="percent of species with genome",x= "")+
  scale_fill_manual(values = class_genome_cols)+
  scale_y_continuous(labels = scales::label_percent())+
  theme_bw()+
  guides(fill="none")

```



```{r}
ggplot(List_of_species)+
  geom_bar(aes(fill=superclass,alpha=genome_match,x=superclass),position=position_fill())+
  coord_flip()+
  labs(y="percent of species with genome",x= "")+
  scale_fill_manual(values = class_col)+
  scale_y_continuous(labels = scales::label_percent())+
  theme_bw()+
  guides(fill="none")
```


```{r}


p.genome<-ggplot(List_of_species)+
  geom_bar(aes(fill=assembly_level,x=superclass))+
  coord_flip()+
  labs(y="Number of Species",x= "",
       title= "Genome availability",
       fill= "assembly level")+
  scale_fill_manual(values = assembly_levels_cols)+
  theme_bw()

p.genome


save_plot(plot = p.genome,base_height = 5,base_width = 8,filename = paste0(path,"/output/figures/p.genome.png"))

```


```{r}
p.genomefibro<-ggplot(List_of_species)+
  geom_bar(aes(x=superclass,fill=cat,alpha=cat_num))+
  coord_flip()+
  labs(y="Number of Species",x= "",
       fill= "group")+
  ggsci::scale_fill_npg()+
  theme_bw()+
  guides(alpha="none")
p.genomefibro

save_plot(plot = p.genomefibro,base_height = 5,base_width = 8,filename = paste0(path,"/output/figures/p.genomefibro.png"))

table(List_of_species$cat,List_of_species$superclass,useNA = "always")
```

# TOGA overlap
```{r}
toga_cols<-c("toga1/2"="#8fbd17","toga1"="#005b88","toga2"="#7852a2","not_in_toga"="grey75")

List_of_species<-List_of_species %>% 
  mutate(TOGA=factor(TOGA_ref,levels=rev(c("toga1/2","toga1","toga2","not_in_toga"))))

TOGA_ol<-ggplot(List_of_species)+
  geom_bar(aes(x=superclass,fill=TOGA))+
  coord_flip()+
  labs(y="Number of Species",x= "",
       fill= "group")+
  scale_fill_manual(values = toga_cols)+
  theme_bw()

TOGA_ol
```


```{r}
TOGA_ol_mam<-List_of_species %>% 
  filter(class== "Mammalia") %>%
  group_by(order) %>% 
  mutate(nspec=n()) %>% 
ggplot()+
  geom_bar(aes(x=reorder(order,nspec),fill=TOGA),show.legend = F)+
  coord_flip()+
  labs(y="Number of Species",x= "",
       fill= "group")+
  scale_fill_manual(values = toga_cols)+
  theme_bw()

TOGA_ol_mam
```

```{r}

annotation_cols <- c(
  "TOGA"                  = "#73CAB2",  # pastel bluish green
  "refseq"                = "#73B1D5",  # pastel blue
  "ensembl"               = "#F1CA73",  # pastel orange
  "TOGA/refseq"           = "#A2D6F3",  # pastel sky blue
  "TOGA/ensembl"          = "#E8A673",  # pastel vermillion
  "refseq/ensembl"        = "#E3B5CF",  # pastel reddish purple
  "TOGA/refseq/ensembl"   = "#737373",  # charcoal (kept darker so it doesn't wash into "no_annotation")
  "no_annotation"         = "grey80"
)

List_of_species<-List_of_species %>% 
  mutate(TOGA= TOGA_species%in%c("toga1/2","toga1","toga2")) %>% 
  mutate(annotation_level = case_when(
    rowSums(across(c(TOGA, refseq, ensembl))) == 0 ~ "no_annotation",
    TRUE ~ pmap_chr(list(TOGA, refseq, ensembl), function(t, r, e) {
      paste(c("TOGA"[t], "refseq"[r], "ensembl"[e]), collapse = "/")
    })
  ))

p.annot<-ggplot(List_of_species)+
  geom_bar(aes(x=superclass,fill=annotation_level))+
  coord_flip()+
  labs(y="Number of Species",x= "",
       fill= "annotation level",
       title = "Genome annotation")+
  scale_fill_manual(values = annotation_cols)+
  theme_bw()

p.annot


library(patchwork)

p.genome / p.annot

save_plot(plot = p.genome / p.annot ,base_height = 5,base_width = 8,filename = paste0(path,"/output/figures/p.genome_assembly_annotation.png"))

```



```{r}
selection<-List_of_species %>% 
  filter(cat=="Both") %>% 
  #filter(superclass != "Fish") %>% 
  mutate(fibro2=if_else(fibro==0,1,as.double(fibro))) %>% 
  mutate(to_process=if_else(fibro2>1,2,as.double(fibro2))) %>% 
  group_by(order) %>% 
  mutate(nspecies=n(),
         order2=if_else(nspecies>2,order,"Other")) %>% 
  group_by(scientific_name) %>% 
  mutate(assembly_level=factor(assembly_level,levels=c("Complete Genome","Chromosome","Scaffold","Contig")))

table(selection$superclass) # species

table(selection$to_process,selection$superclass)


p.genomequality<-selection %>% 
  pivot_longer(cols=c(scaffold_n50,contig_n50), names_to="metric",values_to = "value") %>% 
ggplot()+
  geom_boxplot(aes(x=superclass,y=value,col=assembly_level),outlier.shape = NA)+
  ggbeeswarm::geom_beeswarm(aes(x=superclass,y=value,group=assembly_level),col="grey45",show.legend = T,groupOnX = T,dodge.width = 0.7,alpha=0.5)+
  facet_wrap(.~metric)+
  theme_bw()+
  theme(axis.text.x = element_text(angle=45,hjust=1))+
  scale_y_log10()+
  labs(x="")+
  ggsci::scale_color_futurama()

p.genomequality


save_plot(plot = p.genomequality,base_height = 5,base_width = 5,filename = paste0(path,"/output/figures/p.genomequality.png"))

```


```{r}
p.genome_size<-selection %>% 
  group_by(superclass) %>% 
  mutate(med_g_size=median(genome_size)) %>% 
  ggplot()+
    geom_boxplot(aes(x=superclass,y=genome_size),col="grey75",outlier.shape = NA)+
    ggbeeswarm::geom_beeswarm(aes(x=superclass,y=genome_size,col=superclass),
                              show.legend = T,alpha=0.5)+
    geom_label(aes(x=superclass,y=2*10^9,label=scales::scientific(med_g_size,2),colour=superclass),show.legend = F)+
    theme_bw()+
    theme(axis.text.x = element_text(angle=45,hjust=1))+
    scale_y_log10()+
    labs(x="")+
    scale_fill_manual(values=class_col,limits=force)+
    ggsci::scale_color_jama()

p.genome_size


save_plot(plot = p.genome_size,base_height = 5,base_width = 5,filename = paste0(path,"/output/figures/p.genome_size.png"))
```

Genome size between Mammals and Birds differs by a factor of ~ 2.3. for Chip-seq this might mean we need 2.3 times the number of cells for the same amount of chromatin input into the IP.


```{r}
ggplot(selection) +
  geom_bar(aes(x=order,fill=order),show.legend = F)+
  facet_grid(superclass~.,scales = "free_y",space="free")+
  coord_flip()+
  labs(y="Number of Species")+
  scale_fill_manual(values=order_col)
```

### Select species with best genome and most samples for method establishment
```{r}

best_genome_mammals<-selection %>% 
  filter(superclass=="Mammalia") %>% 
  #group_by(order) %>% 
  slice_max(scaffold_n50,n=30) %>% 
  ungroup() %>% 
  slice_max(fibro>1,n=30) %>% 
  slice_max(individuals,n=20) %>% 
  arrange(fibro)


best_genome_Birds<-selection %>% 
  filter(superclass=="Birds") %>% 
  #group_by(order) %>% 
  ungroup() %>% 
  slice_max(fibro>1,n=10) %>% 
  slice_max(scaffold_n50,n=10)

best_genome<-bind_rows(best_genome_mammals,best_genome_Birds) %>% 
  write_excel_csv(file = paste0(path,"/selection_best_genomes.csv"))

selection_final<-selection %>% 
  filter(superclass!="Fish") %>% 
  filter(!duplicated(scientific_name)) %>% 
  mutate(selected=if_else(species%in%best_genome$species,T,F))

write_excel_csv(selection_final, file = paste0(path,"/selected_species.csv"))

selection_final%>% 
  select(scientific_name,iucn_category,superclass,class,genus,individuals,fibro,tissue_of_origin,assembly_level,scaffold_n50,genome_size,genome_match,selected) %>% 
  write_excel_csv(file = paste0(path,"/selected_species_short.csv"))

```


#Replicates

```{r}


List_of_species %>% 
  mutate(reps=if_else(fibro>2,"3 or more",as.character(fibro))) %>% 
  filter(reps!="0") %>% 
ggplot()+
  geom_bar(aes(x=reps,fill=genome))+
  ggsci::scale_fill_npg()+
  facet_wrap(~superclass)

```

## Sample Sources

```{r}
sample_table<-readRDS(paste0(path,"/output/Sample_List_all_species_clean.rds")) %>% 
  mutate(origin=case_when(origin=="U"~"UNKNOWN",
                          origin=="UNKOWN"~"UNKNOWN",
                          origin=="-"~"UNKNOWN",
                          is.na(origin)~"UNKNOWN",
                          T~origin))

ggplot(sample_table)+
  geom_bar(aes(x=superclass,fill=origin))
```

# Plot the sample provenance
#### make a color scale

from Sample table
```{r}
library(scales)


# Assign categories
df <- tibble(name = sort(unique(sample_table$coming_from_clean))) %>%
  mutate(category = case_when(
    grepl("(?i)aquarium", name) | grepl("(?i)Oceanografic", name) ~ "Aquarium",
    grepl("(?i)zoo", name) ~ "Zoo",
    TRUE ~ "Other"
  ))

# Generate red->orange->yellow palette
zoo_colors <- colorRampPalette(c("red", "orange", "yellow"))(
  length(unique(df$name[df$category == "Zoo"]))
)


# Aquarium colors: blues
aq_colors  <- scales::hue_pal(h = c(200, 260))(
  length(unique(df$name[df$category == "Aquarium"]))
)

# Other colors: greys
oth_colors <- gray.colors(
  length(unique(df$name[df$category == "Other"])), 
  start = 0.2, end = 0.7
)

color_map <- c(
  setNames(zoo_colors, df$name[df$category == "Zoo"]),
  setNames(aq_colors,  df$name[df$category == "Aquarium"]),
  setNames(oth_colors, df$name[df$category == "Other"])
)

```

# plot sample provenance
```{r}
sample_table %>% 
  mutate(superclass=if_else(is.na(superclass),"Sample_discarded",superclass),
         superclass=factor(superclass,levels=rev(c("Mammalia","Birds","Reptil","Amphibia","Fish","Sample_discarded")))) %>% 
  ggplot()+
  geom_bar(aes(x=superclass,fill=coming_from_clean))+
  scale_fill_manual(values= color_map)+
  coord_flip()+
  labs(fill="sample source",x="",y="number of samples")

```

# plot superclass per provider

```{r}
sample_table %>% 
  mutate(superclass=if_else(is.na(superclass),"Sample_discarded",superclass),
         superclass=factor(superclass,levels=rev(c("Mammalia","Birds","Reptil","Amphibia","Fish","Sample_discarded")))) %>% 
  group_by(coming_from_clean) %>% 
  filter(!is.na(superclass)) %>% 
  mutate(samples=n()) %>% 
  ggplot()+
  geom_bar(aes(x=reorder(coming_from_clean,samples),fill=superclass))+
  scale_fill_manual(values = class_col,breaks = names(class_col))+
  coord_flip()+
  labs(fill="class",x="",y="number of samples")

```

# plot fibroblast generation per superclass

```{r}
sample_table %>%
  filter(cell_type!=c("PBMc","unclear")) %>% 
  mutate(fibro_l=factor(if_else(cell_type=="Fibro", "Fibroblast line", "No fibroblast line"),
                        levels=c("No fibroblast line","Fibroblast line")),
         superclass=factor(superclass,
                           levels=rev(c("Mammalia","Birds","Reptil","Amphibia","Fish","Sample_discarded"))
                           )) %>%
  ggplot()+
  geom_bar(aes(x=superclass,fill=factor(fibro_l)),position=position_fill())+
  scale_fill_manual(values = c("grey75","#e63946"),labels=c("No fibroblast line","Fibroblast line"))+
  coord_flip()+
  labs(fill="Fibroblast line established?",x="",y="number of samples")

```


## Save data

```{r}
saveRDS(List_of_species,paste0(path,"/output/Species_List_with_genome_and_cl_info2.rds"))

# write_excel_csv(filter(List_of_species,superclass=="Birds") %>% 
#   ungroup() %>% 
#   select(scientific_name),paste0(path,"/output/bird_names.csv"))


```
