---
title: "2. Plotting the Cryozoo"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

This script summarizes and plots the species currently held in the CryoZoo
collection. It is based on the species table produced by
`Clean_CryoZoo_Lists.Rmd`, so make sure that script has been run first.

## Setup

### Load libraries and colour palettes

```{r,message=F,warning=F}
library(tidyverse)
library(cowplot)
library(readxl)

theme_set(theme_classic())

here::here()

path<-here::here()

# load colour palettes
source("https://raw.githubusercontent.com/mammalian-iPSCs/colour-palettes/main/palettes.R")

```

### Load species table

Quick sanity checks on the species table before any plotting: how many
species already have an established fibroblast line, and how species are
distributed across the five superclasses (Mammalia, Birds, Amphibia,
Reptiles, Fish).

```{r}
List_of_species<-readRDS(paste0(path,"/output/Species_List_with_genome_and_cl_info.rds"))


table(List_of_species$fibro_l)

sum_list<-List_of_species %>%
  group_by(superclass) %>%
  summarize(nspecies=n())

sum(sum_list$nspecies)


table(List_of_species$superclass)

```

## Species overview

### Species counts by class

Overall species counts per superclass, and how many of those already have an
established fibroblast line.

```{r, fig.dim=c(8,7)}

bar.all_spec<-ggplot(data=List_of_species)+
  geom_bar(aes(x="All\n Samples",fill=superclass))+
  geom_text(data=sum_list,aes(label=nspecies),x="All\n Samples",y=c(320,300,255,175,60),color="white")+
  labs(x="",y="Number of species",fill="")+
  coord_flip()+
  theme_minimal()+
  theme(legend.position = "top",
        axis.ticks.y=element_blank(),
        axis.line.y = element_blank())+
  scale_fill_manual(values=class_col,limits=force)



bar.fibro<-ggplot(data=List_of_species)+
  geom_bar(aes(x=superclass,fill=superclass,alpha=fibro_l))+
  labs(x="",y="Number of species",fill="",alpha="Fibroblasts established")+
  coord_flip()+
  ggsci::scale_fill_aaas()+
  theme_minimal()+
  theme(legend.position = c(0.8,0.4),
        axis.ticks.y=element_blank(),
        axis.line.y = element_blank())+
  guides(fill="none")+
  scale_fill_manual(values=class_col,limits=force)


plot_grid(bar.all_spec,bar.fibro,nrow = 2,rel_heights = c(0.3,0.7),align = "v",axis="tblr")


save_plot(plot = plot_grid(bar.all_spec,bar.fibro,nrow = 2,rel_heights = c(0.3,0.7),align = "v",axis="tblr"),base_height = 5,base_width = 8,filename = paste0(path,"/output/figures/overview.png"))



```

### Cell lines per taxonomic order

For each class, species are grouped by order if the order has more than 5
species, otherwise they fall back to a class-level grouping (`taxlevel`).
One bar plot per resulting taxon shows the number of species per family and
whether a fibroblast line exists.

```{r}

taxlevel<-List_of_species %>%
  group_by(order) %>%
  mutate(samples_per_taxlevel=n()) %>%
  group_by(class) %>%
  mutate(taxlevel=if_else(samples_per_taxlevel>5,"order","class"),
        samples_per_taxlevel=if_else(samples_per_taxlevel>5,samples_per_taxlevel,n()),
        taxon=if_else(taxlevel=="order",order,class),
        taxon=if_else(is.na(taxon),superclass,taxon)) %>%
  select(order,class,samples_per_taxlevel,taxlevel,taxon) %>%
  distinct()

table(List_of_species$fibro_l,List_of_species$superclass)

for (i in unique(taxlevel$taxon)) {

print(i)
tl<-taxlevel %>%
  filter(taxon==i) %>%
  select(taxlevel) %>%
  distinct() %>%
  pull(taxlevel)

filt <- List_of_species %>%
  filter(if (tl == "order") order == i else class == i)

filt %>%
ggplot()+
  geom_bar(aes(x=family,fill=superclass,alpha=fibro_l))+
  labs(x="",y="Number of species",fill="",alpha="Fibroblasts?")+
  coord_flip()+
  theme_minimal()+
  theme(legend.position = "right",
        axis.ticks.y=element_blank(),
        axis.line.y = element_blank())+
  guides(fill="none")+
  scale_fill_manual(values=class_col,limits=force)+
  ggtitle(paste(i,"Species with Fibroblast lines"))->p

print(p)
}

```

### IUCN conservation status by class

```{r}


names(IUCN_COLOURS) <- c("LEAST CONCERN", "NEAR THREATENED", "VULNERABLE", "ENDANGERED",
                          "CRITICALLY ENDANGERED", "EXTINCT IN THE WILD", "EXTINCT",
                          "DATA DEFICIENT", "NOT EVALUATED")

bar.iucn<-List_of_species %>%
  filter(!is.na(iucn_category)) %>%
  mutate(iucn_category=factor(iucn_category,levels=c("EXTINCT IN THE WILD","CRITICALLY ENDANGERED","ENDANGERED","VULNERABLE","NEAR THREATENED","LEAST CONCERN","NA")
                              )) %>%
  ggplot()+
  geom_bar(aes(x=superclass,fill=iucn_category),position=position_stack())+
  labs(x="",y="Number of species",fill="IUCN status")+
  coord_flip()+
  theme_minimal()+
  theme(legend.position = c(0.7,0.3),
        axis.ticks.y=element_blank(),
        axis.line.y = element_blank())+
  scale_fill_manual(values=IUCN_COLOURS)
bar.iucn


save_plot(plot = bar.iucn,base_height = 5,base_width = 5,filename = paste0(path,"/output/figures/bar.iucn.png"))

```

### IUCN conservation status overall

Same breakdown as above, collapsed across all classes.

```{r}

List_of_species %>%
  filter(!is.na(iucn_category)) %>%
  mutate(iucn_category=factor(iucn_category,levels=c("EXTINCT IN THE WILD","CRITICALLY ENDANGERED","ENDANGERED","VULNERABLE","NEAR THREATENED","LEAST CONCERN","NA")
                              )) %>%
  ggplot()+
  geom_bar(aes(x="all",fill=iucn_category),position=position_stack())+
  labs(x="",y="Number of species",fill="IUCN status")+
  theme_minimal()+
  scale_fill_manual(values=IUCN_COLOURS)

table(List_of_species$iucn_category)

```

### Replicates per species by class

Distribution of the number of individuals sampled per species, faceted by
class.

```{r,fig.dim=c(10,12)}

p.samples<-ggplot(data=List_of_species)+
  geom_bar(aes(x=individuals,fill=superclass),position = position_dodge(),show.legend = F)+
  labs(x="Number of individuals",y="Number of species",fill="")+
  facet_grid(superclass~.)+
  scale_fill_manual(values=class_col,limits=force)+
  scale_x_continuous(breaks=c(1:20))+
  theme_bw()

p.samples

save_plot(plot = p.samples,base_height = 5,base_width = 5,filename = paste0(path,"/output/figures/p.samples.png"))

```

### Replicates per species, binned

Same information as above, binned into one/two/three/more-than-three
individuals per species for easier comparison across classes.

```{r}

p.samples2<-List_of_species %>%
  mutate(samples_cat=case_when(individuals==1~ "one",
                               individuals==2~ "two",
                               individuals==3 ~ "three",
                               individuals>3 ~ "more than three")) %>%
  mutate(samples_cat=factor(samples_cat,levels=rev(c("more than three","three","two","one")))) %>%
  mutate(superclass=forcats::fct_rev(superclass)) %>%

    ggplot()+
  geom_bar(aes(superclass,fill=samples_cat),position = position_dodge())+
  coord_flip()+
  scale_fill_viridis_d(direction = -1)+
  labs(y="number of species",x="",fill="Replicates per species")+
  theme(legend.position = "top")

p.samples2

save_plot(plot = p.samples2,base_height = 5,base_width = 5,filename = paste0(path,"/output/figures/p.samples2.png"))

```

## Representation relative to global mammal diversity

This section compares CryoZoo's mammal coverage against the full known
mammalian diversity (the "zoonomia" reference group) at the species, genus,
and family level, per order.

### Load global mammal diversity reference data

```{r}

all_species_summary<-readRDS(paste0(path,"/data/mammal_data_orders.rds"))

```

### Species count by order

```{r}
ggplot(all_species_summary)+
  geom_col(aes(y=nspecies,x=group),position = position_dodge())+
  facet_wrap(~order,scales = "free_x")+
  coord_flip()
```

### Genus count by order

```{r}
ggplot(all_species_summary)+
  geom_col(aes(y=ngenera,x=group),position = position_dodge())+
  facet_wrap(~order,scales = "free_x")+
  coord_flip()
```

### Family count by order

```{r}
ggplot(all_species_summary)+
  geom_col(aes(y=nfamily,x=group),position = position_dodge())+
  facet_wrap(~order)+
  coord_flip()
```

### Percent of species covered per order

For each order, the CryoZoo and zoonomia counts are expressed as a percentage
of the total (all-mammal) count, so coverage can be compared directly across
orders of very different sizes.

```{r}
all_mammals<-all_species_summary %>%
  filter(group=="all_mammals") %>%
  ungroup() %>%
  select(-c(group)) %>%
  rename(total_genera=ngenera,
         total_species=nspecies,
         total_families=nfamily)

 all_species_summary<-all_species_summary %>%
   filter(group!="all_mammals") %>%
   left_join(all_mammals) %>%
   mutate(p_species=(nspecies/total_species)*100,
          p_genera=(ngenera/total_genera)*100,
          p_families=(nfamily/total_families)*100)

  ggplot()+
    ggbeeswarm::geom_beeswarm(data=all_species_summary,aes(x=reorder(order,total_species),col=group,y=p_species,size=total_species))+
    coord_flip()+
    theme_light()+
      scale_colour_manual(values=c("zoonomia"="#e63946","cryozoo"="#6987b5"))+
      labs(x="mammalian orders\n (sorted by number of species)",y="percent of all species covered")
```

### Percent covered by taxonomic level

The same coverage comparison, split by taxonomic level (species / genera /
families) and shown side by side for CryoZoo vs. zoonomia.

```{r}
sum_long<-all_species_summary %>%
  pivot_longer(cols=c(p_species,p_genera,p_families), names_to = c("P","tax_level"),values_to = "percent",names_sep = "_") %>%
  select(-P) %>%
  mutate(tax_level=factor(tax_level,levels=rev(c("families","genera","species"))))

ggplot()+
    ggbeeswarm::geom_beeswarm(data=sum_long,aes(x=reorder(order,total_species),colour=group,y=percent),position=position_dodge())+
    coord_flip()+
    theme_light()+
      scale_colour_manual(values=c("zoonomia"="#e63946","cryozoo"="#6987b5"))+
      labs(x="mammalian orders\n (sorted by number of species)",y="percent of all taxonomic units covered") +
  facet_grid(~tax_level)

```

### CryoZoo-only breakdown by taxonomic level

Same as above, restricted to the CryoZoo collection, to see within-order
patterns without the zoonomia reference.

```{r}

taxlevel_colours<- c(
    "species" = "#2AA37F",
    "genera"   = "#E18445",
    "families"  = "#9490C3"
  )

sum_long<-all_species_summary %>%
  filter(group=="cryozoo") %>%
  pivot_longer(cols=c(p_species,p_genera,p_families), names_to = c("P","tax_level"),values_to = "percent",names_sep = "_") %>%
  select(-P) %>%
  mutate(tax_level=factor(tax_level,levels=rev(c("families","genera","species"))))

ggplot()+
    ggbeeswarm::geom_beeswarm(data=sum_long,aes(x=reorder(order,total_species),colour=tax_level,y=percent),position=position_dodge())+
    coord_flip()+
    theme_light()+
      scale_colour_manual(values=taxlevel_colours)+
      labs(x="mammalian orders\n (sorted by number of species)",y="percent of all taxonomic units covered")

```

## Genome sequencing, assembly, and annotation

This section tracks how far each species/order is along the genome pipeline:
sequencing status, assembly quality, and annotation source (TOGA, RefSeq,
Ensembl), alongside fibroblast line availability.

### Reference genome sequencing status by class

```{r}
# per superclass

List_of_species%>%
  mutate(sequencing_status=factor(sequencing_status,levels=rev(c("sample_collected","sample_acquired","in_progress","open","insdc_open","published")))) %>%
  mutate(superclass=factor(superclass,levels=rev(c("Mammalia","Birds","Amphibia","Reptil","Fish")))) %>%

ggplot()+
  geom_bar(aes(fill=sequencing_status,x=superclass))+
  coord_flip()
```

### Reference genome sequencing status by order

```{r}
# per order

List_of_species %>%
  mutate(sequencing_status=factor(sequencing_status,levels=rev(c("sample_collected","sample_acquired","in_progress","open","insdc_open","published")))) %>%
   mutate(superclass=factor(superclass,levels=rev(c("Mammalia","Birds","Amphibia","Reptil","Fish")))) %>%
ggplot()+
  geom_bar(aes(fill=sequencing_status,x=order))+
  coord_flip()+
  facet_grid(superclass~genome,scales = "free",space = "free")



```

### Fibroblast lines established, by order

Restricted to species that already have a fibroblast line: how many species
per order, and how many individuals were sampled per species.

```{r}
List_of_cell_lines<-List_of_species %>%
  filter(fibro_l) %>%
  group_by(order) %>%
  mutate(nspecies=n()) %>%
  ungroup() %>%
  arrange(nspecies)

order_col <- scales::hue_pal()(length(unique(List_of_cell_lines$order)))
names(order_col)<-unique(List_of_cell_lines$order)


p.species_per_order<-ggplot(List_of_cell_lines)+
  geom_bar(aes(y=reorder(order,nspecies),fill=order),show.legend = F)+
  labs(y="",x="Number of species")+
  scale_x_reverse()+
  theme(axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.line.y = element_blank())+
  scale_fill_manual(values=order_col,limits=force)+
  facet_grid(superclass~.,scales = "free",space="free")


p.sample_per_species<-ggplot(List_of_cell_lines)+
  stat_summary(fun.data = "mean_se",aes(x=individuals,y=reorder(order,nspecies),col=order),show.legend = F)+
  #ggbeeswarm::geom_quasirandom(aes(x=individuals,y=reorder(order,nspecies),colour=order),groupOnX = F,show.legend = F)+
  labs(y="",x="Number of samples")+
  scale_x_continuous(breaks=1:15)+
  scale_colour_manual(values=order_col,limits=force)+
  facet_grid(superclass~.,scales = "free",space="free")


p.byorder<-plot_grid(p.species_per_order,p.sample_per_species,rel_widths = c(0.45,0.55))

p.byorder

save_plot(plot = p.byorder,base_height = 8,base_width = 8,filename = paste0(path,"/output/figures/p.byorder.png"))

```

### Genome availability by class

`make_shaded_palette()` takes the existing per-class colours and generates a
lighter/darker shade for each level of a second variable (here: genome
available vs. not), so the "available" bars stay visually tied to their
class colour.

```{r}

library(colorspace)


make_shaded_palette <- function(base_cols, shade_labels, amount = 0.3, name_sep=" ", direction=c("light_to_dark", "dark_to_light")) {
  direction <- match.arg(direction)
  n_shades <- length(shade_labels)

  # spread shade amounts evenly from +amount (lightest) to -amount (darkest)
  # e.g. n=2 -> c(0.3, -0.3); n=3 -> c(0.3, 0, -0.3); n=5 -> c(0.3, 0.15, 0, -0.15, -0.3)
  shade_amounts <- seq(amount, -amount, length.out = n_shades)

  if (direction == "dark_to_light") {
    shade_amounts <- rev(shade_amounts)
  }

  shaded <- lapply(names(base_cols), function(nm) {
    sapply(shade_amounts, function(a) {
      if (a > 0) lighten(base_cols[[nm]], amount = a)
      else if (a < 0) darken(base_cols[[nm]], amount = abs(a))
      else base_cols[[nm]]
    })
  })



  shaded_cols <- unlist(shaded)
  names(shaded_cols) <- as.vector(outer(shade_labels, names(base_cols),
                                         FUN = function(s, n) paste(n, s, sep = name_sep)))

  # reorder so it's grouped by base colour, not by shade
  shaded_cols <- shaded_cols[order(rep(seq_along(base_cols), each = n_shades))]

  shaded_cols
}

class_genome_cols <- make_shaded_palette(class_col,
                                         shade_labels = c("Not Available","Available"),
                                         direction="light_to_dark")

List_of_species %>%
  mutate(genome_class=if_else(genome,paste(superclass, "Available"),paste(superclass,"Not Available")),
         genome_class=factor(genome_class,levels = rev(c("Mammalia Available", "Mammalia Not Available","Birds Available","Birds Not Available","Amphibia Available","Amphibia Not Available","Reptil Available","Reptil Not Available","Fish Available","Fish Not Available")))) %>%
ggplot()+
  geom_bar(aes(fill=genome_class,x=superclass),position=position_fill())+
  coord_flip()+
  labs(y="percent of species with genome",x= "")+
  scale_fill_manual(values = class_genome_cols)+
  scale_y_continuous(labels = scales::label_percent())+
  theme_bw()+
  guides(fill="none")

```

### Genome match: reference vs. sequenced individual

`genome_match` flags whether the available reference genome was sequenced
from the same individual/population represented in the CryoZoo collection.

```{r}
ggplot(List_of_species)+
  geom_bar(aes(fill=superclass,alpha=genome_match,x=superclass),position=position_fill())+
  coord_flip()+
  labs(y="percent of species with genome",x= "")+
  scale_fill_manual(values = class_col)+
  scale_y_continuous(labels = scales::label_percent())+
  theme_bw()+
  guides(fill="none")
```

### Genome assembly level

```{r}


p.genome<-ggplot(List_of_species)+
  geom_bar(aes(fill=assembly_level,x=superclass))+
  coord_flip()+
  labs(y="Number of Species",x= "",
       title= "Genome availability",
       fill= "assembly level")+
  scale_fill_manual(values = assembly_levels_cols)+
  theme_bw()

p.genome


save_plot(plot = p.genome,base_height = 5,base_width = 8,filename = paste0(path,"/output/figures/p.genome.png"))

```

### Genome and fibroblast status combined

`cat` combines genome and fibroblast-line availability into a single status
per species (e.g. "Both", genome only, fibroblast only, neither), used again
later to build the selection shortlist.

```{r}
p.genomefibro<-ggplot(List_of_species)+
  geom_bar(aes(x=superclass,fill=cat,alpha=cat_num))+
  coord_flip()+
  labs(y="Number of Species",x= "",
       fill= "group")+
  ggsci::scale_fill_npg()+
  theme_bw()+
  guides(alpha="none")
p.genomefibro

save_plot(plot = p.genomefibro,base_height = 5,base_width = 8,filename = paste0(path,"/output/figures/p.genomefibro.png"))

table(List_of_species$cat,List_of_species$superclass,useNA = "always")
```

### TOGA annotation overlap, overall

[TOGA](https://github.com/hillerlab/TOGA) provides comparative gene
annotations projected from a reference genome; `TOGA_ref` records whether a
species is covered by the toga1 and/or toga2 annotation runs.

```{r}
toga_cols<-c("toga1/2"="#8fbd17","toga1"="#005b88","toga2"="#7852a2","not_in_toga"="grey75")

List_of_species<-List_of_species %>%
  mutate(TOGA=factor(TOGA_ref,levels=rev(c("toga1/2","toga1","toga2","not_in_toga"))))

TOGA_ol<-ggplot(List_of_species)+
  geom_bar(aes(x=superclass,fill=TOGA))+
  coord_flip()+
  labs(y="Number of Species",x= "",
       fill= "group")+
  scale_fill_manual(values = toga_cols)+
  theme_bw()

TOGA_ol
```

### TOGA annotation overlap, mammals by order

```{r}
TOGA_ol_mam<-List_of_species %>%
  filter(class== "Mammalia") %>%
  group_by(order) %>%
  mutate(nspec=n()) %>%
ggplot()+
  geom_bar(aes(x=reorder(order,nspec),fill=TOGA),show.legend = F)+
  coord_flip()+
  labs(y="Number of Species",x= "",
       fill= "group")+
  scale_fill_manual(values = toga_cols)+
  theme_bw()

TOGA_ol_mam
```

### Combined annotation sources

Species can be annotated by TOGA, RefSeq, and/or Ensembl independently;
`annotation_level` records every combination present so the genome assembly
and annotation views can be stacked into one figure.

```{r}

annotation_cols <- c(
  "TOGA"                  = "#73CAB2",  # pastel bluish green
  "refseq"                = "#73B1D5",  # pastel blue
  "ensembl"               = "#F1CA73",  # pastel orange
  "TOGA/refseq"           = "#A2D6F3",  # pastel sky blue
  "TOGA/ensembl"          = "#E8A673",  # pastel vermillion
  "refseq/ensembl"        = "#E3B5CF",  # pastel reddish purple
  "TOGA/refseq/ensembl"   = "#737373",  # charcoal (kept darker so it doesn't wash into "no_annotation")
  "no_annotation"         = "grey80"
)

List_of_species<-List_of_species %>%
  mutate(TOGA= TOGA_species%in%c("toga1/2","toga1","toga2")) %>%
  mutate(annotation_level = case_when(
    rowSums(across(c(TOGA, refseq, ensembl))) == 0 ~ "no_annotation",
    TRUE ~ pmap_chr(list(TOGA, refseq, ensembl), function(t, r, e) {
      paste(c("TOGA"[t], "refseq"[r], "ensembl"[e]), collapse = "/")
    })
  ))

p.annot<-ggplot(List_of_species)+
  geom_bar(aes(x=superclass,fill=annotation_level))+
  coord_flip()+
  labs(y="Number of Species",x= "",
       fill= "annotation level",
       title = "Genome annotation")+
  scale_fill_manual(values = annotation_cols)+
  theme_bw()

p.annot


library(patchwork)

p.genome / p.annot

save_plot(plot = p.genome / p.annot ,base_height = 5,base_width = 8,filename = paste0(path,"/output/figures/p.genome_assembly_annotation.png"))

```

## Species selection for method establishment

This section narrows the full species list down to a shortlist of
candidates for establishing new methods (e.g. ChIP-seq), based on genome
availability, fibroblast lines, assembly quality, and replicate numbers.

### Candidate pool: genome + fibroblast line available

Only species where `cat == "Both"` (genome and fibroblast line both
available) are considered. `to_process` flags whether more than one
fibroblast line still needs processing.

```{r}
selection<-List_of_species %>%
  filter(cat=="Both") %>%
  #filter(superclass != "Fish") %>%
  mutate(fibro2=if_else(fibro==0,1,as.double(fibro))) %>%
  mutate(to_process=if_else(fibro2>1,2,as.double(fibro2))) %>%
  group_by(order) %>%
  mutate(nspecies=n(),
         order2=if_else(nspecies>2,order,"Other")) %>%
  group_by(scientific_name) %>%
  mutate(assembly_level=factor(assembly_level,levels=c("Complete Genome","Chromosome","Scaffold","Contig")))

table(selection$superclass) # species

table(selection$to_process,selection$superclass)

```

### Genome assembly quality of candidates

Scaffold and contig N50 for the candidate pool, split by class and assembly
level.

```{r}

p.genomequality<-selection %>%
  pivot_longer(cols=c(scaffold_n50,contig_n50), names_to="metric",values_to = "value") %>%
ggplot()+
  geom_boxplot(aes(x=superclass,y=value,col=assembly_level),outlier.shape = NA)+
  ggbeeswarm::geom_beeswarm(aes(x=superclass,y=value,group=assembly_level),col="grey45",show.legend = T,groupOnX = T,dodge.width = 0.7,alpha=0.5)+
  facet_wrap(.~metric)+
  theme_bw()+
  theme(axis.text.x = element_text(angle=45,hjust=1))+
  scale_y_log10()+
  labs(x="")+
  ggsci::scale_color_futurama()

p.genomequality


save_plot(plot = p.genomequality,base_height = 5,base_width = 5,filename = paste0(path,"/output/figures/p.genomequality.png"))

```

### Genome size comparison

```{r}
p.genome_size<-selection %>%
  group_by(superclass) %>%
  mutate(med_g_size=median(genome_size)) %>%
  ggplot()+
    geom_boxplot(aes(x=superclass,y=genome_size),col="grey75",outlier.shape = NA)+
    ggbeeswarm::geom_beeswarm(aes(x=superclass,y=genome_size,col=superclass),
                              show.legend = T,alpha=0.5)+
    geom_label(aes(x=superclass,y=2*10^9,label=scales::scientific(med_g_size,2),colour=superclass),show.legend = F)+
    theme_bw()+
    theme(axis.text.x = element_text(angle=45,hjust=1))+
    scale_y_log10()+
    labs(x="")+
    scale_fill_manual(values=class_col,limits=force)+
    ggsci::scale_color_jama()

p.genome_size


save_plot(plot = p.genome_size,base_height = 5,base_width = 5,filename = paste0(path,"/output/figures/p.genome_size.png"))
```

Genome size between Mammals and Birds differs by a factor of ~ 2.3. for Chip-seq this might mean we need 2.3 times the number of cells for the same amount of chromatin input into the IP.

### Candidate species counts by order

```{r}
ggplot(selection) +
  geom_bar(aes(x=order,fill=order),show.legend = F)+
  facet_grid(superclass~.,scales = "free_y",space="free")+
  coord_flip()+
  labs(y="Number of Species")+
  scale_fill_manual(values=order_col)
```

### Export best-genome shortlist

For mammals and birds separately, rank candidates by assembly quality
(scaffold N50), preference for multiple fibroblast lines, and number of
individuals sampled, then write the shortlist and the full/short selection
tables to CSV.

```{r}

best_genome_mammals<-selection %>%
  filter(superclass=="Mammalia") %>%
  #group_by(order) %>%
  slice_max(scaffold_n50,n=30) %>%
  ungroup() %>%
  slice_max(fibro>1,n=30) %>%
  slice_max(individuals,n=20) %>%
  arrange(fibro)


best_genome_Birds<-selection %>%
  filter(superclass=="Birds") %>%
  #group_by(order) %>%
  ungroup() %>%
  slice_max(fibro>1,n=10) %>%
  slice_max(scaffold_n50,n=10)

best_genome<-bind_rows(best_genome_mammals,best_genome_Birds) %>%
  write_excel_csv(file = paste0(path,"/selection_best_genomes.csv"))

selection_final<-selection %>%
  filter(superclass!="Fish") %>%
  filter(!duplicated(scientific_name)) %>%
  mutate(selected=if_else(species%in%best_genome$species,T,F))

write_excel_csv(selection_final, file = paste0(path,"/selected_species.csv"))

selection_final%>%
  select(scientific_name,iucn_category,superclass,class,genus,individuals,fibro,tissue_of_origin,assembly_level,scaffold_n50,genome_size,genome_match,selected) %>%
  write_excel_csv(file = paste0(path,"/selected_species_short.csv"))

```

## Sample-level summary

The remaining sections switch from the species-level table to the
sample-level table, to look at replicate counts and sample provenance.

### Replicates vs. genome availability

```{r}


List_of_species %>%
  mutate(reps=if_else(fibro>2,"3 or more",as.character(fibro))) %>%
  filter(reps!="0") %>%
ggplot()+
  geom_bar(aes(x=reps,fill=genome))+
  ggsci::scale_fill_npg()+
  facet_wrap(~superclass)

```

### Sample origin by institution type

```{r}
sample_table<-readRDS(paste0(path,"/output/Sample_List_all_species_clean.rds")) %>%
  mutate(origin=case_when(origin=="U"~"UNKNOWN",
                          origin=="UNKOWN"~"UNKNOWN",
                          origin=="-"~"UNKNOWN",
                          is.na(origin)~"UNKNOWN",
                          T~origin))

ggplot(sample_table)+
  geom_bar(aes(x=superclass,fill=origin))
```

### Provenance colour palette

Each providing institution (`coming_from_clean`) is bucketed into Aquarium /
Zoo / Other by name matching, then assigned a colour from a small palette
per bucket (red-orange-yellow for zoos, blues for aquaria, greys for
everything else) so the provenance plots below stay readable with many
distinct providers.

```{r}
library(scales)


# Assign categories
df <- tibble(name = sort(unique(sample_table$coming_from_clean))) %>%
  mutate(category = case_when(
    grepl("(?i)aquarium", name) | grepl("(?i)Oceanografic", name) ~ "Aquarium",
    grepl("(?i)zoo", name) ~ "Zoo",
    TRUE ~ "Other"
  ))

# Generate red->orange->yellow palette
zoo_colors <- colorRampPalette(c("red", "orange", "yellow"))(
  length(unique(df$name[df$category == "Zoo"]))
)


# Aquarium colors: blues
aq_colors  <- scales::hue_pal(h = c(200, 260))(
  length(unique(df$name[df$category == "Aquarium"]))
)

# Other colors: greys
oth_colors <- gray.colors(
  length(unique(df$name[df$category == "Other"])),
  start = 0.2, end = 0.7
)

color_map <- c(
  setNames(zoo_colors, df$name[df$category == "Zoo"]),
  setNames(aq_colors,  df$name[df$category == "Aquarium"]),
  setNames(oth_colors, df$name[df$category == "Other"])
)

```

### Samples by class and provenance

```{r}
sample_table %>%
  mutate(superclass=if_else(is.na(superclass),"Sample_discarded",superclass),
         superclass=factor(superclass,levels=rev(c("Mammalia","Birds","Reptil","Amphibia","Fish","Sample_discarded")))) %>%
  ggplot()+
  geom_bar(aes(x=superclass,fill=coming_from_clean))+
  scale_fill_manual(values= color_map)+
  coord_flip()+
  labs(fill="sample source",x="",y="number of samples")

```

### Class breakdown per provider

```{r}
sample_table %>%
  mutate(superclass=if_else(is.na(superclass),"Sample_discarded",superclass),
         superclass=factor(superclass,levels=rev(c("Mammalia","Birds","Reptil","Amphibia","Fish","Sample_discarded")))) %>%
  group_by(coming_from_clean) %>%
  filter(!is.na(superclass)) %>%
  mutate(samples=n()) %>%
  ggplot()+
  geom_bar(aes(x=reorder(coming_from_clean,samples),fill=superclass))+
  scale_fill_manual(values = class_col,breaks = names(class_col))+
  coord_flip()+
  labs(fill="class",x="",y="number of samples")

```

### Fibroblast establishment rate by class

```{r}
sample_table %>%
  filter(cell_type!=c("PBMc","unclear")) %>%
  mutate(fibro_l=factor(if_else(cell_type=="Fibro", "Fibroblast line", "No fibroblast line"),
                        levels=c("No fibroblast line","Fibroblast line")),
         superclass=factor(superclass,
                           levels=rev(c("Mammalia","Birds","Reptil","Amphibia","Fish","Sample_discarded"))
                           )) %>%
  ggplot()+
  geom_bar(aes(x=superclass,fill=factor(fibro_l)),position=position_fill())+
  scale_fill_manual(values = c("grey75","#e63946"),labels=c("No fibroblast line","Fibroblast line"))+
  coord_flip()+
  labs(fill="Fibroblast line established?",x="",y="number of samples")

```

## Save outputs

```{r}
saveRDS(List_of_species,paste0(path,"/output/Species_List_with_genome_and_cl_info2.rds"))

# write_excel_csv(filter(List_of_species,superclass=="Birds") %>%
#   ungroup() %>%
#   select(scientific_name),paste0(path,"/output/bird_names.csv"))


```
